---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![Travis build status](https://travis-ci.org/ddsjoberg/clintable.svg?branch=master)](https://travis-ci.org/ddsjoberg/clintable)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ddsjoberg/clintable?branch=master&svg=true)](https://ci.appveyor.com/project/ddsjoberg/clintable)
[![Coverage status](https://codecov.io/gh/ddsjoberg/clintable/branch/master/graph/badge.svg)](https://codecov.io/github/ddsjoberg/clintable?branch=master)  
A collection of functions commonly used in the work of the biostatisticians. The goal of **clintable** is to make reporting of tabular analytic results simple, beautiful, and reproducible.  
<!-- Update the list of contributors from the git shell `git shortlog -s -n` -->

## Installation

You can install the production version of **clintable** with:

``` r
install.packages("remotes")
remotes::install_url("https://github.com/ddsjoberg/clintable.git")
```

<!-- and the development version with: -->

<!-- ``` r -->
<!-- install.packages("remotes") -->
<!-- remotes::install_url("https://github.mskcc.org/datadojo/biostatR/archive/development.zip") -->
<!-- ``` -->

## Examples

The vignettes/tutorials for the primary **clintable** functions have detailed examples and can be found at [danieldsjoberg.com/clintable](http://www.danieldsjoberg.com/clintable).  Each vignette is an Rmarkdown file (*.Rmd) and a copy of the files can be found here: [https://github.com/ddsjoberg/clintable/tree/master/vignettes](https://github.com/ddsjoberg/clintable/tree/master/vignettes).  

### Table 1

```{r t1, include = TRUE}
library(clintable)
fmt_table1(trial, by = "trt") %>% 
  add_comparison() %>% 
  bold_labels()

```

### Regression Models

```{r regress, include = TRUE}
mod1 = glm(am ~ mpg + factor(cyl), mtcars, family = binomial(link = "logit"))
fmt_regression(
  mod1, exponentiate = TRUE, 
  label = list(`factor(cyl)` = "No. of Cylinders", mpg = "Miles per Gallon")
  )
```
