---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![Travis build status](https://travis-ci.org/ddsjoberg/gtsummary.svg?branch=master)](https://travis-ci.org/ddsjoberg/gtsummary)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ddsjoberg/gtsummary?branch=master&svg=true)](https://ci.appveyor.com/project/ddsjoberg/gtsummary)
[![Coverage status](https://codecov.io/gh/ddsjoberg/gtsummary/branch/master/graph/badge.svg)](https://codecov.io/github/ddsjoberg/gtsummary?branch=master)  
A collection of functions commonly used in the work of the biostatisticians. The goal of **clintable** is to make reporting of tabular analytic results simple, beautiful, and reproducible.  
<!-- Update the list of contributors from the git shell `git shortlog -s -n` -->

## Notes for next release
- should we change the `id` input for `tbl_summary` to something else like `no_print`.  Currently, we are assuming that any variable listed here is meant (and only meant) to be a grouping variable to calculate p-values for corelated data.  `no_print` would be a more general input to keep variables in `data` but don't print them in summary table.  
- consolidate utility functions under the primary function they are used?  there are so many util files in the R folder  
- look into gt packages fmt_ family of functions.  should we create a function that dictates formatting in our table, and then allows the format to be updated by calls similar to gt package?  
- allow users to pass data frame to regression functions so we can grab the proper labels?  
- currently pseudonyms for functions are NOT supported (i.e. q1 for p35, med for median, etc.)  

## Installation

You can install the production version of **clintable** with:

``` r
install.packages("remotes")
remotes::install_url("https://github.com/ddsjoberg/clintable/archive/master.zip")
```

and the development version with:

``` r
install.packages("remotes")
remotes::install_url("https://github.com/ddsjoberg/clintable/archive/dev.zip")
```

## Examples

The vignettes/tutorials for the primary **clintable** functions have detailed examples and can be found at [danieldsjoberg.com/clintable](http://www.danieldsjoberg.com/clintable).  Each vignette is an Rmarkdown file (*.Rmd) and a copy of the files can be found here: [https://github.com/ddsjoberg/clintable/tree/master/vignettes](https://github.com/ddsjoberg/clintable/tree/master/vignettes).  

### Table 1

```{r t1, include = TRUE}
library(clintable)
fmt_table1(trial, by = "trt") %>% 
  add_comparison() %>% 
  bold_labels()

```

### Regression Models

```{r regress, include = TRUE}
mod1 = glm(am ~ mpg + factor(cyl), mtcars, family = binomial(link = "logit"))
fmt_regression(
  mod1, exponentiate = TRUE, 
  label = list(`factor(cyl)` = "No. of Cylinders", mpg = "Miles per Gallon")
  )
```
