---
title: "Basics of estimated marginal means"
author: "emmeans package, Version `r packageVersion('emmeans')`"
output: emmeans::.emm_vignette
vignette: >
  %\VignetteIndexEntry{Basics of EMMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE, results = "hide", message = FALSE}
require("emmeans")
knitr::opts_chunk$set(fig.width = 4.5, class.output = "ro")
```

<!-- @index Vignettes!Basics -->

## Contents {#contents}

  1. [Motivating example](#motivation)
  2. [EMMs defined](#EMMdef)
      a. [Reference grids](#ref_grid)
      b. [Estimated marginal means](#emmeans)
      c. [Altering the reference grid](#altering)
      d. [Derived covariates](#depcovs)
      d. [Non-predictor variables](#params)
      e. [Graphical displays](#plots)
      f. [Weighting](#weights)
      g. [Multivariate models](#multiv)
  3. [Objects, structures, and methods](#emmobj)
  3. [P values, "significance", and scientific integrity](#pvalues)
  4. [Summary](#summary)
  5. [The back story of **emmeans**](#backstory)
  6. [Further reading](#more)
  
[Index of all vignette topics](vignette-topics.html)


## Why we need EMMs {#motivation}
<!-- @index Examples!`pigs`; Examples!Unbalanced data; 
             Means!Cell; Means!Marginal -->
Consider the `pigs` dataset provided with the package (`help("pigs")` provides
details). These data come from an unbalanced experiment where pigs are given
different percentages of protein (`percent`) from different sources (`source`)
in their diet, and later we measure the concentration (`conc`) of leucine.
Here's an interaction plot showing the mean `conc` at each combination of 
the other factors.
```{r, echo = FALSE}
par(mar = .1 + c(4, 4, 1, 1))   # reduce head space

```
```{r}
with(pigs, interaction.plot(percent, source, conc))
```

This plot suggests that with each `source`, `conc` tends to go up with 
`percent`, but that the mean differs with each `source`.

Now, suppose that we want to assess, numerically, the marginal results for
`percent`.  The natural thing to do is to obtain the marginal means:
```{r}
with(pigs, tapply(conc, percent, mean))
```
Looking at the plot, it seems a bit surprising that the last three means
are all about the same, with the one for 15 percent being the largest.

Hmmmm, so let's try another approach -- actually averaging together the values
we see in the plot. First, we need the means that are shown there:
```{r}
cell.means <- matrix(with(pigs, 
    tapply(conc, interaction(source, percent), mean)), 
    nrow = 3)
cell.means
```
Confirm that the rows of this matrix match the plotted values for fish,
soy, and skim, respectively. Now, average each column:
```{r}
apply(cell.means, 2, mean)
```
These results are decidedly different from the ordinary marginal means we 
obtained earlier. What's going on? The answer is that some observations were
lost, making the data unbalanced:
```{r}
with(pigs, table(source, percent))
```
We can reproduce the marginal means by weighting the cell means with these
frequencies. For example, in the last column:
```{r}
sum(c(3, 1, 1) * cell.means[, 4]) / 5
```
The big discrepancy between the ordinary mean for `percent = 18` and the marginal
mean from `cell.means` is due to the fact that the lowest value receives 3 times
the weight as the other two values.

### The point {#eqwts}
<!-- @index Means!Weighted; Means!Marginal!of cell means@cell
    EMMs!Appropriateness of -->
The point is that the marginal means of `cell.means` give *equal weight* to each
cell. In many situations (especially with experimental data), that is a much
fairer way to compute marginal means, in that they are not biased by imbalances
in the data. We are, in a sense, estimating what the marginal means *would* be,
had the experiment been balanced. Estimated marginal means (EMMs) serve that
need.

All this said, there are certainly situations where equal weighting is *not* 
appropriate. Suppose, for example, we have data on sales of a product given 
different packaging and features. The data could be unbalanced because customers
are more attracted to some combinations than others. If our goal is to 
understand scientifically what packaging and features are inherently more 
profitable, then equally weighted EMMs may be appropriate; but if our goal is to
predict or maximize profit, the ordinary marginal means provide better estimates
of what we can expect in the marketplace.

[Back to Contents](#contents)

## What exactly are EMMs? {#EMMdef}
<!-- @index Estimated marginal means; EMMs; Means!Marginal!Based on a model -->

### Model and reference grid {#ref_grid}
<!-- @index Reference grids; `ref_grid()` -->
Estimated marginal means are based on a *model* -- not directly on data. 
The basis for them is what we call the *reference grid* for a given model.
To obtain the reference grid, consider all the predictors in the model.
Here are the default rules for constructing the reference grid

  * For each predictor that is a *factor*, use its levels (dropping unused ones)
  * For each numeric predictor (covariate), use its average.[^1]
  
The reference grid is then a regular grid of all combinations of these
reference levels.

As a simple example, consider again the `pigs` dataset (see `help("fiber")` for
details). Examination of residual plots from preliminary models suggests that it
is a good idea to work in terms of log concentration.

If we treat the predictor `percent` as a factor, we might fit the 
following model:
```{r}
pigs.lm1 <- lm(log(conc) ~ source + factor(percent), data = pigs)
```
The reference grid for this model can be found via the `ref_grid` function:
```{r}
ref_grid(pigs.lm1)
```
Both predictors are factors, and the reference grid consists of the 
$3\times4 = 12$ combinations of these factor levels. It can be seen explicitly
by looking at the `grid` slot of this object:
```{r}
ref_grid(pigs.lm1) @ grid
``` 
Note that other information is retained in the reference grid, e.g., the
transformation used on the response, and the cell counts as the `.wgt.` column.

Now, suppose instead that we treat `percent` as a numeric predictor. 
This leads to a different model -- and a different reference grid.
```{r}
pigs.lm2 <- lm(log(conc) ~ source + percent, data = pigs)
ref_grid(pigs.lm2)
```
This reference grid has the levels of `source`, but only one `percent` value, 
its average. Thus, the grid has only three elements:
```{r}
ref_grid(pigs.lm2) @ grid
```

[^1]:
In newer versions of **emmeans**, however, covariates having only two distinct values are
by default treated as two-level factors, though there is an option to reduce them to their mean.

[Back to Contents](#contents)

### Estimated marginal means {#emmeans}
<!-- @index Estimated marginal means!Defined; `emmeans()` -->
Once the reference grid is established, we can consider using the model to
estimate the mean at each point in the reference grid. (Curiously, the
convention is to call this "prediction" rather than "estimation"). For
`pigs.lm1`, we have
```{r}
pigs.pred1 <- matrix(predict(ref_grid(pigs.lm1)), nrow = 3)
pigs.pred1
```
Estimated marginal means (EMMs) are defined as equally weighted means of these
predictions at specified margins:
```{r}
apply(pigs.pred1, 1, mean) ### EMMs for source

apply(pigs.pred1, 2, mean) ### EMMs for percent
``` 
For the other model, `pigs.lm2`, we have only one point in the reference
grid for each `source` level; so the EMMs for `source` are just the predictions
themselves:
```{r}
predict(ref_grid(pigs.lm2))
```
These are slightly different from the previous EMMs for `source`, emphasizing
the fact that EMMs are model-dependent. In models with covariates, EMMs are
often called *adjusted means*.

The `emmeans` function computes EMMs, accompanied by standard errors and 
confidence intervals. For example,
```{r}
emmeans(pigs.lm1, "percent")
```

In these examples, all the results are presented on the `log(conc)` scale
(and the annotations in the output warn of this).
It is possible to convert them back to the `conc` scale by back-transforming.
This topic is discussed in [the vignette on transformations](transformations.html).

An additional note: There is an exception to the definition of EMMs given
here. If the model has a nested structure in the fixed effects, then averaging
is performed separately in each nesting group. See the [section on nesting in the
"messy-data" vignette](messy-data.html#nesting) for an example.

[Back to Contents](#contents)

### Altering the reference grid {#altering}
<!-- @index Reference grids!Altering; `ref_grid()`!`cov.reduce`; `ref_grid()`!`cov.keep`; 
            `ref_grid()`!`at`; Examples!`mtcars` -->
It is possible to alter the reference grid. We might, for example, want to 
define a reference grid for `pigs.lm2` that is comparable to the one for 
`pigs.lm1`.
```{r}
ref_grid(pigs.lm2, cov.keep = "percent")
```
Using `cov.keep = "percent"` specifies that, instead of using the mean, the
reference grid should use all the unique values of `each covariate`"percent"`. 

Another option is to specify a `cov.reduce` function that is used in place
of the mean; e.g.,
```{r}
ref_grid(pigs.lm2, cov.reduce = range)
```

Another option is to use the `at` argument. Consider this model for the
built-in `mtcars` dataset:
```{r}
mtcars.lm <- lm(mpg ~ disp * cyl, data = mtcars)
ref_grid(mtcars.lm)
```
Since both predictors are numeric, the default reference grid has only one 
point. For purposes of describing the fitted model, you might want to obtain
predictions at a grid of points, like this:
```{r}
mtcars.rg <- ref_grid(mtcars.lm, cov.keep = 3,
                      at = list(disp = c(100, 200, 300)))
mtcars.rg
```
This illustrates two things: a new use of `cov.keep` and the `at` argument.
`cov.keep = "3"` specifies that any covariates having 3 or fewer unique values
is treated like a factor (the system default is `cov.keep = "2"`). 
The `at` specification gives three values of `disp`,
overriding the default behavior to use the mean of `disp`. 
Another use of `at` is to focus on only some of the levels of a factor. Note that
`at` does not need to specify every predictor; those not mentioned in `at` are
handled by `cov.reduce`, `cov.keep`, or the default methods. Also, covariate values
in `at` need not be values that actually occur in the data, whereas `cov.keep`
will use only values that are achieved.

[Back to Contents](#contents)


### Derived covariates {#depcovs}
<!-- @index Covariates!Derived; Quadratic terms; Polynomial regression -->
You need to be careful when one covariate depends on the value of another. To
illustrate in the `mtcars` example, suppose we want to use `cyl` as a factor and
include a quadratic term for `disp`:
```{r}
mtcars.1 <- lm(mpg ~ factor(cyl) + disp + I(disp^2), data = mtcars)
emmeans(mtcars.1, "cyl")
```
Some users may not like function calls in the model formula, so they instead do something like this:
```{r}
mtcars <- transform(mtcars, 
                    Cyl = factor(cyl),
                    dispsq = disp^2)
mtcars.2 <- lm(mpg ~ Cyl + disp + dispsq, data = mtcars)
emmeans(mtcars.2, "Cyl")
```
Wow! Those are really different results -- even though the models are
equivalent. Why is this? To understand, look at the reference grids:
```{r}
ref_grid(mtcars.1)
ref_grid(mtcars.2)
```
For both models, the reference grid uses the `disp` mean of 230.72. But for
`mtcars.2`, we also set `dispsq` to its mean of 68113. This is not right,
because `dispsq` should be the square of `disp` (about 53232, not 68113) in order to be
consistent. If we use that value of `dispsq`, we get the same results (modulus
rounding error) as for `mtcars.1`:
```{r}
emmeans(mtcars.2, "Cyl", at = list(dispsq = 230.72^2))
```

In summary, for polynomial models and others where some covariates depend on
others in nonlinear ways, include that dependence in the model formula (as in
`mtcars.1`) using `I()` or `poly()` expressions, or alter the reference grid so
that the dependency among covariates is correct.


### Non-predictor variables {#params}
<!-- @index `params`; Variables that are not predictors; -->
Reference grids are derived using the variables in the right-hand side of the model
formula. But sometimes, these variables are not actually predictors. For example:
```{r, eval = FALSE}
deg <- 2
mod <- lm(y ~ treat * poly(x, degree = deg), data = mydata)
```
If we call `ref_grid()` or `emmeans()` with this model, it will try to construct
a grid of values of `treat`, `x`, and `deg` -- causing an error because `deg` is
not a predictor in this model. To get things to work correctly, you need to name
`deg` in a `params` argument, e.g.,
```{r, eval = FALSE}
emmeans(mod, ~ treat | x, at = list(x = 1:3), params = "deg")
```


[Back to Contents](#contents)

### Graphical displays {#plots}
<!-- @index Graphical displays; Plots!of EMMs@emms; Plots!Interaction-style; `emmip()`;   -->
The results of `ref_grid()` or `emmeans()` (these are objects of class `emmGrid`)
may be plotted in two different 
ways. One is an interaction-style plot, using `emmip()`. In the following, let's 
use it to compare the predictions from `pigs.lm1` and `pigs.lm2`:
```{r}
emmip(pigs.lm1, source ~ percent)
emmip(ref_grid(pigs.lm2, cov.reduce = FALSE), source ~ percent)
```

Notice that `emmip()` may also be used on a fitted model. The formula
specification needs the *x* variable on the right-hand side and the "trace"
factor (what is used to define the different curves) on the left.
This is a good time to yet again emphasize that EMMs are based on a *model*.
Neither of these plots is an interaction plot of the *data*; they are 
interaction plots of model predictions; and since both models do not include
an interaction, no interaction at all is evident in the plots.

###### {#plot.emmGrid}
<!-- @index Plots!of confidence intervals@conf; `plot.emmGrid()` -->
The other graphics option offered is the `plot()` method for `emmGrid` objects. In
the following, we display the estimates and 95% confidence intervals for
`mtcars.rg` in separate panels for each `disp`.
```{r}
plot(mtcars.rg, by = "disp")
```

This plot illustrates, as much as anything else, how silly it is to try to
predict mileage for a 4-cylinder car having high displacement, or an 8-cylinder
car having low displacement. The widths of the intervals give us a clue that we
are extrapolating. A better idea is to acknowledge that displacement largely
depends on the number of cylinders. So here is yet another way to 
use `cov.reduce` to modify the reference grid:
```{r}
mtcars.rg_d.c <- ref_grid(mtcars.lm, at = list(cyl = c(4,6,8)),
                          cov.reduce = disp ~ cyl)
mtcars.rg_d.c @ grid
```
The `ref_grid` call specifies that `disp` depends on `cyl`; so a linear model 
is fitted with the given formula and its fitted values are used as the `disp`
values -- only one for each `cyl`. If we plot this grid, the results are 
sensible, reflecting what the model predicts for typical cars with each 
number of cylinders:
```{r fig.height = 1.5}
plot(mtcars.rg_d.c)
```

###### {#ggplot}
<!-- @index **ggplot2** package -->
Wizards with the **ggplot2** package can further enhance these plots if 
they like. For example, we can add the data to an interaction plot -- this
time we opt to include confidence intervals and put the three sources 
in separate panels:
```{r}
require("ggplot2")
emmip(pigs.lm1, ~ percent | source, CIs = TRUE) +
    geom_point(aes(x = percent, y = log(conc)), data = pigs, pch = 2, color = "blue")
```

[Back to Contents](#contents)

### Using weights {#weights}
<!-- @index Means!Weighted; `emmeans()`!`weights` -->
It is possible to override the equal-weighting method for computing EMMs. Using
`weights = "cells"` in the call will weight the predictions according to their
cell frequencies (recall this information is retained in the reference grid).
This produces results comparable to ordinary marginal means:
```{r}
emmeans(pigs.lm1, "percent", weights = "cells")
```
Note that, as in the ordinary means in [the motivating example](#motivation),
the highest estimate is for `percent = 15` rather than `percent = 18`. It is
interesting to compare this with the results for a model that includes only
`percent` as a predictor.
```{r}
pigs.lm3 <- lm(log(conc) ~ factor(percent), data = pigs)
emmeans(pigs.lm3, "percent")
```
The EMMs in these two tables are identical, but their standard errors are
considerably different. That is because the model `pigs.lm1` accounts for 
variations due to `source`. The lesson here is that it is possible to obtain 
statistics comparable to ordinary marginal means, while still accounting for
variations due to the factors that are being averaged over.

[Back to Contents](#contents)

### Multivariate responses {#multiv}
<!-- @index Multivariate models; Examples!`MOats`
     Examples!Multivariate; `ref_grid()`!`mult.name` -->
The **emmeans** package supports various multivariate models. When there
is a multivariate response, the dimensions of that response are treated as if
they were levels of a factor. For example, the `MOats` dataset provided in the
package has predictors `Block` and `Variety`, and a four-dimensional response
`yield` giving yields observed with varying amounts of nitrogen added to the soil.
Here is a model and reference grid:
```{r}
MOats.lm <- lm (yield ~ Block + Variety, data = MOats)
ref_grid (MOats.lm, mult.name = "nitro")
```
So, `nitro` is regarded as a factor having 4 levels corresponding to the 4
dimensions of `yield`. We can subsequently obtain EMMs for any of the factors
`Block`, `Variety`, `nitro`, or combinations thereof. The argument `mult.name =
"nitro"` is optional; if it had been excluded, the multivariate levels would
have been named `rep.meas`.

[Back to Contents](#contents)





## Objects, structures, and methods {#emmobj}
<!-- @index `emmGrid` objects -->
The `ref_grid()` and `emmeans()` functions are introduced previously.
These functions, and a few related ones,
return an object of class `emmGrid`:
```{r}
pigs.rg <- ref_grid(pigs.lm1)
class(pigs.rg)

pigs.emm.s <- emmeans(pigs.rg, "source")
class(pigs.emm.s)
```
If you simply show these objects, you get different-looking results:
```{r}
pigs.rg

pigs.emm.s
```
This is based on guessing what users most need to see when displaying the object.
You can override these defaults; for example to just see a quick summary of
what is there, do
```{r}
str(pigs.emm.s)
```
<!-- @index `str()`; `summary()`; `summary_emm` object; `print.summary_emm()` -->
The most important method for `emmGrid` objects is `summary()`. It is used as the
print method for displaying an `emmeans()` result. For this reason, arguments for 
`summary()` may also be specified within most functions that produce `these kinds of
results.`emmGrid` objects. For example:
```{r}
# equivalent to summary(emmeans(pigs.lm1, "percent"), level = 0.90, infer = TRUE))
emmeans(pigs.lm1, "percent", level = 0.90, infer = TRUE)
```


This `summary()`
method for `emmGrid` objects) actually produces a `data.frame`, but with extra bells 
and whistles:
```{r}
class(summary(pigs.emm.s))
```
This can be useful to know because if you want to actually *use* `emmeans()` results
in other computations, you should save its summary, and then you can access those
results just like you would access data in a data frame. The `emmGrid` object itself
is not so accessible. There is a `print.summary_emm()` function that is what
actually produces the output you see above -- a data frame with extra
annotations.

[Back to Contents](#contents)




## P values, "significance", and scientific integrity {#pvalues}
<!-- @index Significance!Assessing; 
    "Significant"@significant!Why you shouldn't say this;
    *P* values!Interpreting; ATOM; 
    ASA Statement on *P* values -->
There is a growing consensus among statisticians and researchers that *P* values
are widely misunderstood, and that the term "statistical significance" is
downright misleading. If you have a small *P* value, it *only* means that the
effect being tested is unlikely to be explained by chance variation alone, in
the context of the current study and the current statistical model underlying
the test. If you have a large *P* value, it *only* means that the observed
effect could plausibly be due to chance alone: it is *wrong* to conclude that
there is no effect.

The American Statistical Association has for some time been advocating very
cautious use of *P* values (Wasserman *et al.* 2014, 2019). The ASA now
*advises against ever using the term "statistically significant"* 
(or a related descriptor) because it is
too often misinterpreted, and too often used carelessly. 

Wasserman *et al.* (2019),
and the 43 articles it accompanies in the same issue of *TAS*, recommend a
number of alternatives. I do not agree with all that is said in the main
article, and there are portions that are too cutesy or wander off-topic.
Further, it is quite dizzying to try to digest all the accompanying articles,
and to reconcile their disagreeing viewpoints. But the basic ideas are
summarized with the help of the acronym **ATOM**, for the four highlighted words
below:

  * **Acceptance** of uncertainty. If you
    were to repeat your study, you would get somewhat different results and perhaps
    different effects will stand out. Likewise, there are other reasonable
    models that you might have chosen, and those would have yielded different
    results. If the topic is worth studying, others will try to replicate it,
    perhaps with different study designs and different ways of operationalizing
    important concepts -- again, yielding different results.
    
  * **Thoughtfulness**. Design your study with clear objectives in mind, and
    spend time considering the best way to address those objectives in your study
    design. Pilot the study to ensure that the procedures actually work, how much
    variability to expect, and ensure that the sample size is adequate. How do
    your results fit in the big picture, scientifically? Think about the
    importance of the effects you have observed in your model(s)/data
    in the context of the science. Think carefully about how well your model 
    fits the data and if there are other possibilities.
    
  * **Openness**. Be transparent in disclosing the methods used, and reporting 
    comprehensively all the successes and failures, the small and large effects.
    If hypothesis tests are used, report the actual *P* values found (e.g., "*P* = 0.075")
    and don't suppress or withhold results just because they are "nonsignificant,"
    because those results may still be useful to other researchers via meta-analysis.
    Trust in the expert judgment of other researchers.
    
  * **Modesty**. Be clear in examining and expressing the limitations of your work.
    Understand that other researchers with different subject-area strengths may choose
    other models that may cast a different light on your work. Encourage others to try to
    reproduce your work, and be a good scientific colleague to them. Do not
    exaggerate the importance of your findings. Your study is not final or
    definitive; recognize that it is just one part of a broader picture.
    Ironically, the more important the research, the more related research will
    be undertaken, making that picture bigger (and, potentially,  your part of it smaller).

  

### References
 Wasserman RL, Lazar NA (2016)
   "The ASA's Statement on *p*-Values: Context, Process, and Purpose,"
   *The American Statistician*, **70**, 129--133,
   https://doi.org/10.1080/00031305.2016.1154108
 
 Wasserman RL, Schirm AL, Lazar, NA (2019)
   "Moving to a World Beyond 'p < 0.05',"
   *The American Statistician*, **73**, 1--19,
   https://doi.org/10.1080/00031305.2019.1583913




[Back to Contents](#contents)


## Summary of main points {#summary}
  * EMMs are derived from a *model*. A different model for the same data may lead
    to different EMMs.
  * EMMs are based on a *reference grid* consisting of all combinations
    of factor levels, with each covariate set to its average (by default).
  * For purposes of defining the reference grid, dimensions of
    a multivariate response are treated as levels of a factor.
  * EMMs are then predictions on this reference grid, or marginal
    averages thereof (equally weighted by default).
  * Reference grids may be modified using `at` or `cov.reduce`;
    the latter may be logical, a function, or a formula.
  * Reference grids and `emmeans()` results may be plotted via `plot()`
    (for parallel confidence intervals) or `emmip()` (for an interaction-style
    plot).
  * Be cautious with the terms "significant" and "nonsignificant", and avoid using hard
    thresholds like *P* < 0.05 as a criterion for scientific meaning or
    importance. The importance of a result is established by scientific context,
    not by some *P* value. Remember ATOM.
    
[Back to Contents](#contents)


## The back story of **emmeans** {#backstory}
<!-- @index Back story of **emmeans**; University administration; **lsmeans** -->
The **emmeans** package and its predecessor, **lsmeans**, were developed in part
because I wanted it for teaching. This is hardly a surprise, as I am an
academic. I had taught experimental design and analysis a number of times,
usually requiring SAS to do the kind of *post hoc* comparisons that I like to
encourage people to do. SAS is fine software (in spite of the opinions of some
Rusers with attitude), but it is quite different from R, and more and more, I
had students who were bewildered by it. Some colleagues of mine think it is
absolutely essential that masters-level students learn to use SAS. I can see
that point, but my job was to teach design, not SAS -- and there are a lot 
of topics to cover.

But here's what you wouldn't expect: the real impetus to develop the package
came from doing some consulting work for the university administration. They
wanted help in analyzing faculty salaries, especially with regard to gender and
ethnicity differences. Accordingly, we fitted a pretty complicated regression
model with effects for gender and ethnicity, as well as such things as college
department, rank, tenure status, seniority, etc. It is a really messy model.
I first came to the conclusion that to make sense of the model results
in terms that the administration could understand, I could obtain predictions 
from that model and average them together in meaningful ways to summarize the 
primary factor effects. I figured out how to write R code to estimate those
averages and test differences among them. Only after all that did I
realize that really what I was computing was least-squares means (or EMMs);
so the package was born.

It grew from there, especially in terms of supporting more models, evolving from
a spaghetti-code design to one incorporating essential building blocks with
methods for different models. Meanwhile, I retired, so I had time on my hands to
dink around with it (and to keep learning about new or unfamiliar models
and data-analysis approaches). 

One aspect of the package that I am especially proud of is the ability to
express results in accessible ways when there is a transformed response or link
function (see the ["transformations" vignette](transformations.html)). Again,
this grew out of my needs for that administrative salary analysis. The model
that fitted the best used a reciprocal transformation, but I wanted to express
the predicted salaries in dollars. Moreover, I wanted to compare them using
ratios. Those requirements led directly to the development of the `regrid()`
function and the options to metamorphose any `emmGrid` object to the response or
the log scale.

The lesson is that it is possible that even university service duties can lead
to technical developments in statistics. One never knows where new ideas and
inspirations will arise. Speaking of this, I have learned and benefited
enormously from users' questions about the package, and basically all recent
improvements to **emmeans** are a direct result of those communications. So
don't be shy about asking questions or requesting features.

[Back to Contents](#contents)


## Further reading {#more}
The reader is referred to other vignettes for more details and advanced use.
The strings linked below are the names of the vignettes; i.e., they can
also be accessed via `vignette("`*name*`", "emmeans")`

  * Models that are supported in **emmeans** (there are lots of them)
    ["models"](models.html)
  * Confidence intervals and tests: 
    ["confidence-intervals"](confidence-intervals.html)
  * Often, users want to compare or contrast EMMs: ["comparisons"](comparisons.html)
  * Working with response transformations and link functions:
    ["transformations"](transformations.html)
  * Multi-factor models with interactions: ["interactions"](interactions.html)
  * Working with messy data and nested effects: ["messy-data"](messy-data.html)
  * Making predictions from your model: ["predictions"](predictions.html)
  * Examples of more sophisticated models (e.g., mixed, ordinal, MCMC)
    ["sophisticated"](sophisticated.html)
  * Utilities for working with `emmGrid` objects: ["utilities"](utilities.html)
  * Frequently asked questions: ["FAQs"](FAQs.html)
  * Adding **emmeans** support to your package: ["xtending"](xtending.html)

[Back to Contents](#contents)

[Index of all vignette topics](vignette-topics.html)