---
title: "Confidence intervals and tests in emmeans"
author: "emmeans package, Version `r packageVersion('emmeans')`"
output: emmeans::.emm_vignette
vignette: >
  %\VignetteIndexEntry{Confidence intervals and tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE, results = "hide", message = FALSE}
require("emmeans")
knitr::opts_chunk$set(fig.width = 4.5, class.output = "ro")
```

<!-- @index Vignettes!Confidence intervals and tests -->

## Contents {#contents}
This vignette describes various ways of summarizing `emmGrid` objects.

  1. [`summary()`, `confint()`, and `test()`](#summary)
  2. [Back-transforming to response scale](#tran)
     (See also the ["transformations" vignette](transformations.html))
  3. [Multiplicity adjustments](#adjust)
  4. [Using "by" variables](#byvars)
  5. [Joint (omnibus) tests](#joint)
  6. [Testing equivalence, noninferiority, nonsuperiority](#equiv)
  7. Graphics (in ["basics" vignette](basics.html#plots))
  
[Index of all vignette topics](vignette-topics.html)

## `summary()`, `confint()`, and `test()` {#summary}
<!-- @index `summary()`; `summary()`!`infer`; `confint()`; `test()`; 
    Confidence intervals; `summary()`!Calculated columns; 
    `summary()`!Show sample size; Sample size, displaying;
            Tests!One- and two-sided; Tests!Nonzero null;
            `summary()`!Bayesian models; `summary()`!`hpd.summary()`;
            Examples!`pigs` -->
The most important method for `emmGrid` objects is `summary()`. For one thing, it is
called by default when you display an `emmeans()` result.
The `summary()` function has a lot of options, and the detailed documentation via 
`help("summary.emmGrid")` is worth a look.

For ongoing illustrations, let's re-create some of the objects in the ["basics" vignette](basics.html) for the `pigs` example:
```{r}
pigs.lm1 <- lm(log(conc) ~ source + factor(percent), data = pigs)
pigs.rg <- ref_grid(pigs.lm1)
pigs.emm.s <- emmeans(pigs.rg, "source")
```

Just `summary(<object>)` by itself will produce a summary that varies somewhat
according to context. It does this by setting different defaults for the `infer`
argument, which consists of two logical values, specifying confidence intervals 
and tests, respectively. [The exception is models fitted using MCMC methods,
where `summary()` is diverted to the `hpd.summary()` function, a 
preferable summary for many Bayesians.]

The summary of a newly made reference grid will show just estimates and
standard errors, but not confidence intervals or tests (that is, `infer =
c(FALSE, FALSE)`). The summary of an `emmeans()` result, as we see above, will
have intervals, but no tests (i.e., `infer = c(TRUE, FALSE)`); and the result of
a `contrast()` call (see [comparisons and contrasts](comparisons.html)) will
show test statistics and *P* values, but not intervals (i.e., `infer = c(FALSE,
TRUE)`). There are courtesy methods `confint()` and `test()` that just call
`summary()` with the appropriate `infer` setting; for example,
```{r}
test(pigs.emm.s)
```
It is not particularly useful, though, to test these EMMs against the default of 
zero -- which is why tests are not usually shown. It makes a lot more sense
to test them against some target concentration, say 40. And suppose we want to do a one-sided test to see if the concentration is greater than 40. Remembering that the
response is log-transformed in this model, 
```{r}
test(pigs.emm.s, null = log(40), side = ">")
```
It is also possible to add calculated columns to the summary, via the `calc` 
argument. The calculations can include any columns up through `df` in the summary,
as well as any variable in the object's `grid` slot. Among the latter are 
usually weights in a column named `.wgt.`, and we can use that to include
sample size in the summary:
```{r}
confint(pigs.emm.s, calc = c(n = ~.wgt.))
```


[Back to Contents](#contents)


## Back-transforming {#tran}
<!-- @index Transformations!Back-transforming; `type`; Response scale -->
Transformations and link functions are supported an several ways in **emmeans**,
making this a complex topic worthy of [its own vignette](transformations.html).
Here, we show just the most basic approach. Namely, specifying the argument
`type = "response"` will cause the displayed results to be back-transformed
to the response scale, when a transformation or link function is incorporated
in the model. For example, let's try the preceding `test()` call again:
```{r}
test(pigs.emm.s, null = log(40), side = ">", type = "response")
```
Note what changes and what doesn't change. In the `test()` call, we *still* use
the log of 40 as the null value; `null` must always be specified on the
linear-prediction scale, in this case the log. In the output, the displayed
estimates, as well as the `null` value, are shown back-transformed. As well, the
standard errors are altered (using the delta method). However, the *t* ratios
and *P* values are identical to the preceding results. That is, the tests
themselves are still conducted on the linear-predictor scale (as is noted in the
output).

Similar statements apply to confidence intervals on the response scale:
```{r}
confint(pigs.emm.s, side = ">", level = .90, type = "response")
```
With `side = ">"`, a *lower* confidence limit is computed on the log scale, then
that limit is back-transformed to the response scale. (We have also illustrated 
how to change the confidence level.)

[Back to Contents](#contents)


## Multiplicity adjustments {#adjust}
<!-- @index Multiplicity adjustments; `adjust`; 
     Tukey adjustment; Sidak adjustment -->
Both tests and confidence intervals may be adjusted for simultaneous inference. 
Such adjustments ensure that the confidence coefficient for a whole set of 
intervals is at least the specified level, or to control for multiplicity
in a whole family of tests. This is done via the `adjust`
argument. For `ref_grid()` and `emmeans()` results, the default is `adjust =
"none"`. For most `contrast()` results, `adjust` is often something else,
depending on what type of contrasts are created. For example, pairwise
comparisons default to `adjust = "tukey"`, i.e., the Tukey HSD method.
The `summary()` function sometimes *changes* `adjust` if it is inappropriate.
For example, with
```{r}
confint(pigs.emm.s, adjust = "tukey")
```
the adjustment is changed to the Sidak method because the Tukey adjustment is inappropriate unless you are doing pairwise comparisons.

####### {#adjmore}
<!-- @index **mvtnorm** package; Bonferroni adjustment
     Multivariate *t* (`"mvt"`) adjustment; Unadjusted tests -->
An adjustment method that is usually appropriate is Bonferroni; however, it can
be quite conservative. Using `adjust = "mvt"` is the closest to being the
"exact" all-around method "single-step" method, as it uses the multivariate *t*
distribution (and the **mvtnorm** package) with the same covariance structure as
the estimates to determine the adjustment. However, this comes at high
computational expense as the computations are done using simulation techniques.
For a large set of tests (and especially confidence intervals), the
computational lag becomes noticeable if not intolerable.

For tests, `adjust` increases the *P* values over those otherwise obtained with
`adjust = "none"`. Compare the following adjusted tests with the unadjusted ones
previously computed.
```{r}
test(pigs.emm.s, null = log(40), side = ">", adjust = "bonferroni")
```

[Back to Contents](#contents)


## "By" variables {#byvars}
<!-- @index `by` groups; Grouping into separate sets -->
Sometimes you want to break a summary down into smaller pieces; for this
purpose, the `by` argument in `summary()` is useful. For example,
```{r}
confint(pigs.rg, by = "source")
```
If there is also an `adjust` in force when `by` variables are used, the 
adjustment is made *separately* on each `by` group; e.g., in the above, we would
be adjusting for sets of 4 intervals, not all 12 together.

There can be a `by` specification in `emmeans()` (or equivalently, a `|` in the
formula); and if so, it is passed on to `summary()` and used unless overridden
by another `by`. Here are examples, not run:
```{r eval = FALSE}
emmeans(pigs.lm, ~ percent | source)     ### same results as above
summary(.Last.value, by = percent)       ### grouped the other way
```
Specifying `by = NULL` will remove all grouping.

### Simple comparisons {#simple}
<!-- @index Simple comparisons; `contrast()`!`simple`; `simple = "each"` -->
There is also a `simple` argument for `contrast()` that is in essence 
the inverse of `by`; the contrasts are run using everything *except* the
specified variables as `by` variables. To illustrate, let's consider 
the model for `pigs` that includes the interaction (so that the levels
of one factor compare differently at levels of the other factor).
```{r}
pigsint.lm <- lm(log(conc) ~ source * factor(percent), data = pigs)
pigsint.rg <- ref_grid(pigsint.lm)
contrast(pigsint.rg, "consec", simple = "percent")
```
In fact, we may do *all* one-factor comparisons by specifying `simple = "each"`.
This typically produces a lot of output, so use it with care.

[Back to Contents](#contents)


## Joint tests {#joint}
<!-- @index `test()`!`joint = TRUE`; `joint`; Type III tests -->
From the above, we already know how to test individual results. For pairwise comparisons (details in [the "comparisons" vignette](comparisons.html)), we might do
```{r}
pigs.prs.s <- pairs(pigs.emm.s)
pigs.prs.s
```
But suppose we want an *omnibus* test that all these comparisons are zero.
Easy enough, using the `joint` argument in `test` (note: the `joint` argument
is *not* available in `summary()`; only in `test()`):
```{r}
test(pigs.prs.s, joint = TRUE)
```
Notice that there are three comparisons, but only 2 d.f. for the test,
as cautioned in the message.

The test produced with `joint = TRUE` is a "type III" test (assuming the default
equal weights are used to obtain the EMMs). See more on these types of tests for
higher-order effects in the ["interactions" vignette section on
contrasts](interactions.html#contrasts).

####### {#joint_tests}
<!-- @index `joint_tests()`; Type III tests; Analysis of variance!Type III -->
For convenience, there is also a `joint_tests()` function that performs
joint tests of contrasts among each term in a model or `emmGrid` object.
```{r}
joint_tests(pigsint.rg)
```
The tests of main effects are of families of contrasts; those for interaction
effects are for interaction contrasts. These results are essentially the same as
a "Type-III ANOVA", but may differ in situations where there are empty cells or
other non-estimability issues, or if generalizations are present such as unequal
weighting. (Another distinction is that sums of squares and mean squares are not
shown; that is because these really are tests of contrasts among predictions,
and they may or may not correspond to model sums of squares.) 

One may use `by` variables with `joint_tests`. For example:
```{r}
joint_tests(pigsint.rg, by = "source")
```
In some models, it is possible to specify `submodel = "type2"`, thereby obtaining 
something akin to a Type II analysis of variance. See the [messy-data vignette](messy-data.html#type2submodel) for an example.


[Back to Contents](#contents)


## Testing equivalence, noninferiority, and nonsuperiority {#equiv}
<!-- @index Tests!Equivalence; Tests!Noninferiority; `test()`!`delta` -->
The `delta` argument in `summary()` or `test()` allows the user to
specify a threshold value to use in a test of equivalence, noninferiority, 
or nonsuperiority. An equivalence test is kind of a backwards significance 
test, where small *P* values are associated with small differences relative 
to a specified threshold value `delta`.
The help page for `summary.emmGrid` gives the details of 
these tests.
Suppose in the present example, we consider two sources to be equivalent if they 
are within 25% of each other. We can test this as follows:
```{r}
test(pigs.prs.s, delta = log(1.25), adjust = "none")
```
By our 25% standard, the *P* value is quite small for comparing soy and skim, 
providing statistical evidence that their difference is enough smaller than the
threshold to consider them equivalent.

[Back to Contents](#contents)


## Graphics {#graphics}
Graphical displays of `emmGrid` objects are described in the 
["basics" vignette](basics.html#plots)

[Index of all vignette topics](vignette-topics.html)
