---
title: "Global Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{global_options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

There are various global options that can be set for {gtsummary}.
Here we review a complete list of options available to users.

## Styling Estimates

The {gtsummary} includes various default functions for styling and formatting statistics.
These defaults can be modified with global options.

### p-values

P-values are reported throughout {gtsummary} tables.
The default function to style p-values is `style_pvalue()`.
Large p-values are rounded to one decimal place in `style_pvalue()`.
Use the `gtsummary.pvalue_fun` option to change the function that styles p-values.
To report large p-values rounded to two decimal places rather that one, change the default function to `style_pvalue(x, digits = 2)`.

```{r}
options(gtsummary.pvalue_fun = function(x) style_pvalue(x, digits = 2))
```

Or equivalently
 
```{r}
options(gtsummary.pvalue_fun = purrr::partial(style_pvalue, digits = 2))
``` 

### percentages

In `tbl_summary()`, the default function for styling percentages is `style_percent()`.
If you'd like to update the default, for example, because you'd prefer percentages rounded to one decimal place, you may include the `gtsummary.tbl_summary.percent_fun` option.

```{r}
options(gtsummary.pvalue_fun = function(x) sprintf(x * 100, fmt='%#.1f'))
```

### survival estimates

Similar to the above options, the default function for styling estimates in `tbl_survival()` can be modified with `gtsummary.tbl_survival.estimate_fun`.

## Format {gt} Tables

Tables created with the {gt} package are incredibly customizable.
You can set an option to apply additional {gt} formatting to all {gtsummary} tables.
Every table returned from {gtsummary} contains a list of {gt} calls that are applied when printed.  For example, here are the first few associated with a `tbl_summary()` table.

```{r, eval = TRUE}
library(gtsummary)
tbl_summary(trial)$gt_calls %>% head(n = 4)
```

You may append {gt} calls to this list using the `gtsummary.as_gt.addl_cmds` option---with the following caveats:

1. Only include {gt} functions that apply to *all* tables. For example, a {gt} function that changes the font size would be appropriate.

1. All {gt} functions must be referenced using the double-colon prefix---`gt::foo()`. These functions will execute when the {gt} library is not loaded.

1. Do not include the `data=` argument in any {gt} function call. The function will appear in a chained sequence of {gt} calls with the `%>%` operator.  The data argument is taken from the previous function in the sequence.

In the example below, the font size and cell padding is reduced.

```{r}
options(gtsummary.as_gt.addl_cmds = "gt::tab_options(table.font.size = 'small', row.padding = gt::px(1))")
```


## Printing

{gtsummary} uses the {gt} package to print all summary tables.
In addition to supporting {gt}, the {gtsummary} package works well with `knitr::kable()`.
This is particularly useful when outputting documents to Microsoft Word.
If the {gt} package is not installed, {gtsummary} will fall back to `knitr::kable()`.
To explicitly set the printing engine, set the option in the script or in the user- or project R profile, `.Rprofile`.

```{r}
options(gtsummary.print_engine = "kable") 
```

or

```{r}
options(gtsummary.print_engine = "gt")
```

Output from {kable} is less full featured compared to summary tables produced with {gt}.
For example, {kable} summary tables do not include indentation, footnotes, and spanning header rows.
