---
title: "Global Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Global Options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
library(gtsummary)
```

There are various global options that can be set for {gtsummary}.
Here we review a complete list of options available to users.

## Styling Estimates

The {gtsummary} package includes various default functions for styling and formatting statistics.
These defaults can be modified with global options.

### p-values

P-values are reported throughout {gtsummary} tables.
The default function to style p-values is `style_pvalue()`, where large p-values are rounded to one decimal place.
Use the `gtsummary.pvalue_fun` option to change the function that styles p-values.
For example, to report large p-values rounded to two decimal places rather that one, change the default function to `style_pvalue(x, digits = 2)`.

```{r}
options(gtsummary.pvalue_fun = function(x) style_pvalue(x, digits = 2))
```

### percentages

In `tbl_summary()`, the default function for styling percentages is `style_percent()`.
If you'd like to update the default, for example, because you'd prefer percentages rounded to one decimal place, you may include the `gtsummary.tbl_summary.percent_fun` option.

```{r}
options(gtsummary.tbl_summary.percent_fun = function(x) sprintf(x * 100, fmt='%#.1f'))
```

### survival estimates

Similar to the above options, the default function for styling estimates in `tbl_survival()` can be modified with `gtsummary.tbl_survival.estimate_fun`.

## Printing

{gtsummary} uses the [{gt}](https://gt.rstudio.com/) package to print all summary tables.
In addition to supporting {gt}, the {gtsummary} package works well with [`knitr::kable()`](https://rdrr.io/cran/knitr/man/kable.html).
This is particularly useful when outputting documents to Microsoft Word.
If the {gt} package is not installed, {gtsummary} will fall back to `knitr::kable()`.
To explicitly set the printing engine, set the option in the script or in the user- or project R profile, `.Rprofile`.

```{r}
options(gtsummary.print_engine = "kable") 
```

or

```{r}
options(gtsummary.print_engine = "gt")
```

Output from {gt} is more full-featured compared to a summary table produced with {kable}.
For example, {gt} summary tables can include indentation, footnotes, and spanning header rows.

## Format {gt} Tables

Tables created with the [{gt}](https://gt.rstudio.com/) package are incredibly customizable.
You can set an option to apply additional {gt} formatting to all {gtsummary} tables.
Every table returned from {gtsummary} maybe converted to a {gt} object with `as_gt()`.

You may append additional [{gt} calls](https://gt.rstudio.com/reference/index.html) the `gtsummary.as_gt.addl_cmds` option---with the following caveats:

1. Only include {gt} functions you want to apply to *all* tables. For example, a {gt} function that changes the font size would be appropriate. A table-specific header may not.

1. All {gt} functions must be referenced using the double-colon prefix (i.e. `gt::foo()`). These functions will execute when the {gt} library is not loaded.

1. Do not include the `data=` argument in any {gt} function call. The function will appear in a chained sequence of {gt} calls with the `%>%` operator.  The data argument is taken from the previous function in the sequence.

In the example below, the font size and cell padding are reduced globally.

```{r}
options(gtsummary.as_gt.addl_cmds = "gt::tab_options(table.font.size = 'small', data_row.padding = gt::px(1))")
```

## Default tests in `add_p()`

To update the default tests in the `add_p()` function, you can set the following global options.

```{r, echo=FALSE, eval = TRUE}
tibble::tribble(
  ~`Option Name`, ~`Tests Modified`,
  "gtsummary.add_p.test.continuous_by2",           "continuous variables with 2-level by variable",
  "gtsummary.add_p.test.continuous",               "continuous variables with 3- or more level by variable",
  "gtsummary.add_p.test.categorical",              "categorical/dichotomous variables",
  "gtsummary.add_p.test.categorical.low_count",    "categorical/dichotomous variables with minimum expected count <5 in one cell",
  "gtsummary.add_p.test.categorical.group_by2",    "categorical/dichotomous grouped/correlated variables with 2-level by variable",
  "gtsummary.add_p.test.continuous.group_by2",     "continuous grouped/correlated variables with 2-level by variable"
) %>%
  gt::gt()
```

For example, to report the t-test and one-way ANOVA for continuous variables, use the following options:

```{r}
options(
  gtsummary.add_p.test.continuous_by2 = "t.test",
  gtsummary.add_p.test.continuous = "aov"
)
```
