---
title: "gtsummary + R markdown"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gtsummary_with_rmarkdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Tables printed with {gtsummary} can be seamlessly integrated into R markdown documents.
The {gtsummary} package was written to be a companion to the [{gt} package](https://gt.rstudio.com/) from RStudio, and {gtsummary} tables are printed using {gt} when possible.
Currently, {gt} supports **HTML** output, with **LaTeX** and **RTF** planned for the future.

### R Markdown Output Types

- `html_document`
  - HTML output is ully supported by {gtsummary} and is the recommended output type.
- `pdf_document`
  - PDF for {gt} is currently under development, and {gtsummary} output will be printed with the `knitr::kable()` function.
  - You can force printing with {gt} by including `options(gtsummary.print_engine = "gt")` at the top of your script, or a converting a {gtsummary} object to {gt} with `as_gt()`, e.g. `tbl_summary(trial) %>% as_gt()`. 
  - If you print to PDF using {gt} you may get a gorgeous table, but you may also get an error or a malformed table.
- `rtf_document`
  - You may print to Microsoft Word via RTF; however, RTF output is under development in the {gt} package. You may get a gorgeous table, but you may also get an error or a malformed table. 
  - `kintr::kable()` does not support RTF output.
  - To avoid any messaging about RTF format being under development, include `options(gtsummary.print_engine = "gt")` at the top of your script, or convert {gtsummary} objects to {gt} with `as_gt()`, e.g. `tbl_summary(trial) %>% as_gt()`. 
- `word_document`
  - If the specified output is `word_document`, `knitr::kable()` will be used to print the table. 
  - {gt} does not support word document output.

When a table is printed with `kintr::kable()` the resulting table is less full featured compared to a table printed with {gt}.
For example, the table will not contain footnotes, spanning headers, or row indentation.

### Example R Markdown Report

An example R markdown report using {gtsummary} has been included with the package.
To open the example file, run the following command in the R console.

```{r, eval = FALSE}
file.path(system.file(package = "gtsummary"), 
          "rmarkdown_example", "gtsummary_rmarkdown_html.Rmd") %>%
  file.edit()
```
