---
title: "Tutorial: tbl_summary"
date: "Last Updated: January 13, 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Tutorial: tbl_summary}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `tbl_summary()` function calculates **descriptive statistics** for continuous, categorical, and dichotomous variables in **R**, and presents the results in a **beautiful, customizable summary table** perfect for creating tables ready for publication (for example, Table 1 or demographic tables).

This vignette will walk a reader through the `tbl_summary()` function, and the various functions available to modify and make additions to an existing table summary object.

To start, a quick note on the {magrittr} package's pipe function, `%>%`.  By default the pipe operator puts whatever is on the left hand side of `%>%` into the first argument of the function on the right hand side.  The pipe function can be used to make the code relating to `tbl_summary()` easier to use, but it is *not* required.  Here are a few examples of how `%>%` translates into typical R notation.

    x %>% f() is equivalent to f(x)
    x %>% f(y) is equivalent to f(x, y)
    y %>% f(x, .) is equivalent to f(x, y)
    z %>% f(x, y, arg = .) is equivalent to f(x, y, arg = z)

Here's how this translates into the use of `tbl_summary()`.

    mtcars %>% tbl_summary() is equivalent to tbl_summary(mtcars)
    mtcars %>% tbl_summary(by = am) is equivalent to tbl_summary(mtcars, by = am)
    tbl_summary(mtcars, by = am) %>% add_p() is equivalent to
        tbl = tbl_summary(mtcars, by = am)
        add_p(tbl)
    
## Setup

Before going through the tutorial, install {gtsummary} and {gt}.

```{r, include=FALSE}
library(gtsummary)
library(dplyr)
```

```{r, eval=FALSE}
install.packages("gtsummary")
remotes::install_github("rstudio/gt", ref = gtsummary::gt_sha)

library(gtsummary)
library(dplyr)
```

- If you experience issues installing {gt} on Windows, install [Rtools from CRAN](https://cran.r-project.org/bin/windows/Rtools/), restart R, and attempt installation again.

## Example data set

We'll be using the [`trial`](http://www.danieldsjoberg.com/gtsummary/reference/trial.html) data set throughout this example.

* This set contains data from `r nrow(trial)` patients who received one of two types of chemotherapy (Drug A or Drug B).
The outcomes are tumor response and death.  

* Each variable in the data frame has been assigned an **attribute label** (i.e. `attr(trial$trt, "label") == "Chemotherapy Treatment")` with the **[labelled](http://larmarange.github.io/labelled/)** package, which we highly recommend using. These labels are displayed in the {gtsummary} output table by default. Using {gtsummary} on a data frame without labels will simply print variable names, or there is an option to add labels later.  


        trt      Chemotherapy Treatment
        age      Age, yrs
        marker   Marker Level, ng/mL
        stage    T Stage
        grade    Grade
        response Tumor Response
        death    Patient Died
        ttdeath  Years from Treatment to Death/Censor  

* Our example dataset has a mix of continuous, dichotomous (0/1), and categorical variables, some with missing data (NA). 

```{r, message=FALSE}
head(trial)
```

For brevity in the tutorial, let's keep a subset of the variables from the trial data set.

```{r}
trial2 =
  trial %>%
  dplyr::select(trt, marker, stage)
```

## Basic Usage

The default output from `tbl_summary()` is meant to be **publication ready**. 

Let's start by creating a table of summary statistics from the `trial` data set.  The `tbl_summary()` function can take, at minimum, a data frame as the only input, and returns descriptive statistics for each column in the data frame.


```{r, message=FALSE}
tbl_summary(trial2)
```

Note the sensible defaults with this basic usage (that can be customized later):

* **Variable types** are automatically detected so that appropriate descriptive statistics are calculated.

* **Label attributes** from the dataset are automatically printed.

* **Missing values** are listed as "Unknown" in the table.

* Variable levels are **indented** and **footnotes** are added if printed using [{gt}](https://gt.rstudio.com/). (can alternatively be printed using `knitr::kable()`; see options [here](http://www.danieldsjoberg.com/gtsummary/articles/global_options.html#printing))

This is a great basic table, but for this study data the summary statistics should be **split by treatment group**, which can be done by using the **`by = `** argument.
To compare two or more groups, include [`add_p()`](http://www.danieldsjoberg.com/gtsummary/reference/add_p.html) with the function call, which detects variable type and uses an appropriate test.

```{r}
tbl_summary(trial2, by = trt) %>% add_p()
```

## Customize Output

There are four primary ways to customize the output of the summary table.

1. Modify `tbl_summary()` function input arguments
1. Add additional data/information to a summary table with `add_*()` functions
1. Modify summary table appearance with the {gtsummary} functions
1. Modify table appearance with {gt} package functions

### Modifying `tbl_summary()` function arguments

The `tbl_summary()` function includes many input options for modifying the appearance.

    label       specify the variable labels printed in table  
    type        specify the variable type (e.g. continuous, categorical, etc.)
    statistic   change the summary statistics presented  
    digits      number of digits the summary statistics will be rounded to  
    missing     whether to display a row with the number of missing observations 
    sort        change the sorting of categorical levels by frequency
    percent     print column, row, or cell percentages

### {gtsummary} functions to add information

The {gtsummary} package has built-in functions for adding to results from `tbl_summary()`.  The following functions add columns and/or information to the summary table.

    add_p()           add p-values to the output comparing values across groups   
    add_overall()     add a column with overall summary statistics   
    add_n()           add a column with N (or N missing) for each variable   
    add_stat_label()  add a column showing a label for the summary statistics shown in each row   
    add_q()           add a column of q values to control for multiple comparisons   

### {gtsummary} functions to format table

The {gtsummary} package comes with functions specifically made to modify and format summary tables.  

    modify_header()         relabel columns in summary table  
    bold_labels()           bold variable labels  
    bold_levels()           bold variable levels  
    italicize_labels()      italicize variable labels  
    italicize_levels()      italicize variable levels  
    bold_p()                bold significant p-values  

### {gt} functions to format table

The {gt} package is packed with many great functions for modifying table output---too many to list here.  Review the package's website for a full listing. https://gt.rstudio.com/index.html

To use the {gt} package functions with {gtsummary} tables, the summary table must first be converted into a `gt` object.  To this end, use the `as_gt()` function after modifications have been completed with {gtsummary} functions.

```{r, eval = FALSE}
trial %>%
  tbl_summary(by = trt, missing = "no") %>%
  add_n() %>%
  as_gt() %>%
  <gt functions>
```

### Example

The code below calculates the standard table with summary statistics split by treatment with the following modifications

- Report 'mean (SD)' and 'n / N (\%)'  
- Round the marker mean and SD to 1 and 2 places, respectively
- Modify variable labels in the table 
- Use t-test instead of Wilcoxon rank-sum 
- Round large p-values to two decimal place  
- Add column with statistic labels  
- Modify header to include percentages in each group  
- Bold variable labels  
- Italicize variable levels  

```{r}
trial2 %>%
  # build base summary table
  tbl_summary(
    # split table by treatment variable
    by = trt,
    # change variable labels
    label = list(marker ~ "Marker, ng/mL",
                 stage ~ "Clinical T Stage"),
    # change statistics printed in table
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} / {N} ({p}%)"),
    digits = list("marker" ~ c(1, 2))
  ) %>%
  # add p-values, report t-test, round large pvalues to two decimal place
  add_p(test = list(marker ~ "t.test"),
                 pvalue_fun = function(x) style_pvalue(x, digits = 2)) %>%
  # add statistic labels
  add_stat_label() %>%
  # bold variable labels, italicize levels
  bold_labels() %>%
  italicize_levels() %>%
  # bold p-values under a given threshold (default is 0.05)
  bold_p(t = 0.2) %>%
  # include percent in headers
  modify_header(stat_by = "**{level}**, N = {n} ({style_percent(p, symbol = TRUE)})")
```

Each of the modification functions have additional options outlined in their respective help files.

## Select Helpers {#select_helpers}

There is flexibility in how you select variables for {gtsummary} arguments, which allows for many customization opportunities! For example, if you want to show age and the marker levels to one decimal place in `tbl_summary()`, you can pass `digits = c(age, marker) ~ 1`. The selecting input is flexible, and you may also pass quoted column names.

Going beyond typing out specific variables in your dataset, you can use: 

1. All [**{tidyselect} helpers**](https://tidyselect.r-lib.org/reference/select_helpers.html) available throughout the tidyverse, such as `starts_with()`, `contains()`, and `everything()` (i.e. anything you can use with the `dplyr::select()` function can be used with {gtsummary}).


1. Additional [**{gtsummary} selectors**](http://www.danieldsjoberg.com/gtsummary/reference/select_helpers.html) that are included in the package to supplement tidyselect functions.

    + **Summary type** There are three types of summary types in {gtsummary}, and you may use the type to select columns.  This is useful, for example, when you wish to report the mean and standard deviation for all continuous variables: `statistic = all_continuous() ~ "{mean} ({sd})"`.

      ```{r, eval=FALSE}
      all_continuous()      all_categorical()      all_dichotomous()
      ```

    + **Vector class or type** Select columns based on their class or type.

      ```{r, eval=FALSE}
      all_numeric()         all_integer()          all_logical() 
      all_factor()          all_character()        all_double()
      ```

Most variable selecting can be accomplished in a single R object (e.g. `c(age, stage)`).
However, you may need more complex compound select structure, which cannot be represented by a single object.
For example, if you want to select all columns except `"stage"`, you'll need to utilize the `vars(...)` function.
When the `vars(...)` function is used, all arguments from `vars()` are passed directly to `dplyr::select()`.
`vars(everything(), -stage)` is the `dplyr` equivalent of `select(everything(), -stage)`.

### Examples

In the example below, we report the mean and standard deviation for continuous variables, and percent for all categorical.  We'll report t-tests rather than Wilcoxon rank-sum test for continuous variables, and report Fisher's exact test for response.

Note that dichotomous variables are, by default, included with `all_categorical()`.  Use `all_categorical(dichotomous = FALSE)` to exclude dichotomous variables.

```{r}
trial %>%
  dplyr::select(trt, response, age, stage, marker, grade) %>%
  tbl_summary(
    by = trt,
    type = list(c(response, grade) ~ "categorical"), # select by variables in c()
    statistic = list(all_continuous() ~ "{mean} ({sd})", 
                     all_categorical() ~ "{p}%") # select by summary type
  ) %>%
  add_p(test = list(contains("response") ~ "fisher.test", # select using functions in tidyselect
                    all_continuous() ~ "t.test"))
```


## Advanced Customization {#advanced}

When you print output from the `tbl_summary()` function into the R console or into an R markdown, there are default printing functions that are called in the background: `print.tbl_summary()` and `knit_print.tbl_summary()`.  The true output from `tbl_summary()` is a named list, but when you print the object, a formatted version of `.$table_body` is displayed.  All formatting and modifications are made using the {gt} package.

```{r}
tbl_summary(trial2) %>% names()
```

These are the additional data stored in the `tbl_summary()` output list. 

    table_body   data frame with summary statistics  
    meta_data    data frame that is one row per variable with data about each  
    by, df_by    the by variable name, and a  data frame with information about the by variable  
    call_list    named list of each function called on the `tbl_summary` object  
    inputs       inputs from the `tbl_summary()` function call  
    gt_calls     named list of {gt} function calls  
    kable_calls  named list of function calls to be applied before knitr::kable()  

`gt_calls` is a named list of saved {gt} function calls.  The {gt} calls are run when the object is printed to the console or in an R markdown document. Here's an example of the first few calls saved with `tbl_summary()`:

```{r}
tbl_summary(trial2) %>% purrr::pluck("gt_calls") %>% head(n = 5)
```

The {gt} functions are called in the order they appear, always beginning with the `gt()` function.

If the user does not want a specific {gt} function to run (i.e. would like to change default printing), any {gt} call can be excluded in the `as_gt()` function.
For example, the `tbl_summary()` call creates many named {gt} function calls: *`r tbl_summary(trial2) %>% purrr::pluck("gt_calls") %>% names() %>% paste(collapse = ", ")`*.
Any of these can be excluded.  In the example below, the default footnote will be excluded from the output.

After the `as_gt()` function is run, additional formatting may be added to the table using {gt} formatting functions.
In the example below, a spanning header for the `by = ` variable is included with the {gt} function `tab_spanner()`.

```{r as_gt2, eval=FALSE}
tbl_summary(trial2, by = trt) %>%
  as_gt(include = -tab_footnote) %>%
  gt::tab_spanner(label = gt::md("**Treatment Group**"),
                  columns = gt::starts_with("stat_"))
```

```{r as_gt1, echo=FALSE}
# this code chunk only works if gt is installed
if (requireNamespace("gt", quietly = TRUE)) {
  tbl_summary(trial2, by = trt) %>%
    as_gt(include = -tab_footnote) %>%
    gt::tab_spanner(label = gt::md("**Treatment Group**"),
                    columns = gt::starts_with("stat_"))
}
```

## Setting Default Options {#options}

The `tbl_summary()` function and it's related functions have sensible defaults for rounding and formatting results.  If you, however, would like to change the defaults there are a few options.  The default options can be changed for a single script with addition an `options()` command in the script.  The defaults can also be set on the project- or user-level R profile, `.Rprofile`.

```{r, eval=FALSE}
help("Rprofile")

usethis::edit_r_profile()
```

The following parameters are available to be set:

```{r, echo=FALSE}
data.frame(
  `Description` = c("Formatting and rounding p-values", 
                    "Formatting and rounding percentages",
                    "Print tables with `gt` or `kable`"),
  `Example` = c("`options(gtsummary.pvalue_fun = function(x) gtsummary::style_pvalue(x, digits = 2))`",
                '`options(gtsummary.tbl_summary.percent_fun = function(x) sprintf("%.2f", 100 * x))`',
                '`options(gtsummary.print_engine = "kable")`   `options(gtsummary.print_engine = "gt")`'),
  `Functions` = c("`add_p()`, `tbl_regression()`, `tbl_uvregression()`",
                  "`tbl_summary()`",
                  "All `tbl_*()` functions")
) %>% 
  knitr::kable()
```

When setting default rounding/formatting functions, set the default to a function object rather than an evaluated function.  For example, if you want to round percentages to 3 significant figures use,

```{r,eval=FALSE}
options(gtsummary.tbl_summary.percent_fun = function(x) sigfig(x, digits = 3))
```

