---
title: "Tutorial: tbl_regression"
author: "Daniel D. Sjoberg"
date: "Last Updated: November 29, 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: tbl_regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
library(gtsummary)
library(dplyr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette will walk a reader through the `tbl_regression()` function, and the various functions available to modify and make additions to an existing formatted regression table. `tbl_regression()` uses `broom::tidy()` to perform the initial model formatting, and can accommodate many different model types (e.g. `lm()`, `glm()`, `coxph()`, `survreg()` and more). If your class of model is not supported by `broom::tidy()`, please request support at github.com/ddsjoberg/gtsummary/issues.  In most cases, it is simple to support a new class of model.

To start, a quick note on the `magrittr` package's pipe function, `%>%`.  By default the pipe operator puts whatever is on the left hand side of `%>%` into the first argument of the function on the right hand side.  The pipe function can be used to make the code relating to `tbl_regression()` easier to use, but it is *not* required.  Here are a few examples of how `%>%` translates into typical R notation.

    x %>% f() is equivalent to f(x)
    x %>% f(y) is equivalent to f(x, y)
    y %>% f(x, .) is equivalent to f(x, y)
    z %>% f(x, y, arg = .) is equivalent to f(x, y, arg = z)

## Basic Usage

We will use the `trial` data set throughout this example.  This set contains data from `r nrow(trial)` patients randomized to a new adjuvant treatment or placebo.  The outcome is a binary tumor response.  Each variable in the data frame has been assigned an attribute label (i.e. `attr(trial$trt, "label") == "Treatment Randomization")`

    trt      Treatment Randomization
    age      Age, yrs
    marker   Marker Level, ng/mL
    stage    T Stage
    grade    Grade
    response Tumor Response

The default output from `tbl_regression()` is meant to be publication ready.  Let's start by creating a regression model table from the `trial` data set included in the `gtsummary` package.  We will predict tumor response using age, stage, and grade using a logistic regression model.  

```{r, message=FALSE}
# build logistic regression model
m1 = glm(response ~ age + stage + grade, trial, family = binomial(link = "logit"))

# format results into data frame
tbl_regression(m1, exponentiate = TRUE)
```

Because the variables in the data set were labeled, the labels were carried through into the model regression results table.  (Had the data not been labeled, the default is to display the variable name.)  The model was recognized as logistic regression with coefficients exponentiated, so the header displayed "OR" for odds ratio. The correct reference group has also been added to the table.

Imagine you are not interested in the individual p-values for the stage and grade levels.  You can use the `add_global()` function to add a global p-value for each.  The `add_global()` function is a wrapper for `cars::Anova()`, which calculates LRT and Wald test statistics (a few other methods are also available depending on the type of model). The p-value presented is the default from `cars::Anova()` for your type of model.  You can pass arguments to this function via `add_global()` to change the defaults.

```{r, message=FALSE}
# format results into data frame with global p-values
tbl_regression(m1, exponentiate = TRUE) %>% 
  add_global()
```

## Customize Regression Table Output

There are four primary was to customize the output of the regression model table.

1. Modify `tbl_regression()` function input arguments
1. Add additional data/information to a summary table with `add_*()` functions
1. Modify summary table appearance with the gtsummary functions
1. Modify table appearance with gt package functions

### Modifying function arguments

The `tbl_regression()` function includes many input options for modifying the appearance.

  > `label` modify the variable labels printed in the table  
  > `exponentiate` exponentiriate model coefficients
  > `include` names of variables to include in output. Default is all variables.
  > `show_yesno` Show both levels fo yes/no variables. Default prints on single row.
  > `conf.level` confidence leve of confidence interval  
  > `intercept` logical argument indicates whether to include the intercept in output
  > `coef_fun` function to round and format beta coefficients  
  > `pvalue_fun` function to round and format p-values  

### Functions to add information

The gtsummary package has built-in functions for adding to results from `tbl_regression()`.  The following functions add columns and/or information to the regression table.

  > `add_global()` Adds the global p-value for a categorical variables   

### gtsummary functions to format table

The gtsummary package comes with functions specifically made to modify and format summary tables.  These functions share the same prefixes as functions from the gt package to ease use.

  > `tab_style_bold_labels()` Bold variable labels  
  > `tab_style_bold_levels()` Bold variable levels  
  > `tab_style_italicize_labels()` Italicize variable labels  
  > `tab_style_italicize_levels()` Italicize variable levels  
  > `tab_style_bold_p()` Bold significant p-values  

### gt functions to format table

The gt package is packed with many great functions for modifying table output---too many to list here.  Review the package's website for a full listing. https://gt.rstudio.com/index.html

To use the gt package functions with gtsummary tables, the regression table must first be converted into a `gt` object.  To this end, use the `as_gt()` function after modifications have been completed with gtsummary functions.

```
m1 %>%
  tbl_regression(exponentiate = TRUE) %>%
  <gt functions>
```
### Example

There are formatting options available, such as bolding and italicizing rows.  In the example below,  

1. Variable labels are bold  
1. Levels of categorical levels are italicized  
1. P-values less than 0.10 are bold  
1. Large p-values are rounded to two decimal places  
1. Coefficients are exponentiated to give odds ratios  
1. Coefficients are rounded to 3 significant figures

```{r, message=FALSE}
# format results into data frame with global p-values
m1 %>%
  tbl_regression( 
    exponentiate = TRUE, 
    pvalue_fun = function(x) style_pvalue(x, digits = 2),
    coef_fun = function(x) style_sigfig(x, digits = 3)
  )  %>% 
  add_global() %>%
  tab_style_bold_p(t = 0.10) %>%
  tab_style_bold_labels() %>% 
  tab_style_italicize_levels()
```

##  Report Results Inline

Reproducible reports are an import part of good practices.  We often need to report the results from a table in the text of an R markdown report.  Inline reporting has been made simple with `inline_text()`.

Let's first create a regression model table.

```{r}
tbl_m1 = tbl_regression(m1, exponentiate = TRUE)
tbl_m1
```

To report the result for `age`, use the following commands inline.

> `` `r
inline_text(tbl_m1, variable = "age")` `` 

Here's how the line will appear in your report.

> `r inline_text(tbl_m1, variable = "age")`

It is reasonable that you'll need to modify the text.  To do this, use the `pattern` argument.  The `pattern` argument syntax follows the `glue::glue()` format with referenced R objects being inserted between curly brackets.  You have access to following fields within the `pattern` argument.  The default is `pattern = "{coef} ({conf.level*100}% CI {ll}, {ul}; {pvalue})"`

    {coef}       Primary estimate (e.g. odds ratio, model coefficient)
    {ll}         Rounded lower limit of confidence interval
    {ul}         Rounded upper limit of confidence interval
    {pvalue}     Rounded p-value
    {conf.level} Confidence level of interval

The `inline_text` function has has arguments for rounding the p-value (`pvalue_fun`) and the coefficients and confidence interval (`coef_fun`).

> Age was not significantly associated with tumor response `` `r
inline_text(tbl_m1, variable = "age", pattern = "(OR {coef}; 95% CI {ll}, {ul}; {pvalue})")` ``.

> Age was not significantly associated with tumor response `r inline_text(tbl_m1, variable = "age", pattern = "(OR {coef}; 95% CI {ll}, {ul}; {pvalue})")`.

If you're printing results from a categorical variable, include `level` argument, e.g. `inline_text(tbl_m1, variable = "stage", level = "T3")` resolves to "`r inline_text(tbl_m1, variable = "stage", level = "T3")`".  


## Advanced Customization

When you print the output from the `tbl_regression()` function into the R console or into an R markdown, there are default printing functions that are called in the background: `print.tbl_regression()` and `knit_print.tbl_regression()`.  The true output from `tbl_regression()` is a named list, but when you print the object, a formatted version of `.$table_body` is displayed.  All formatting and modifications are made using the `gt` package.

```{r}
tbl_regression(m1) %>% names()
```

These are the additional data stored in the `tbl_regression()` output list. 

  > `table_body` Data frame with summary statistics  
  > `n` N included in model  
  > `model_obj` the model object passed to `tbl_regression`  
  > `call_list` Named list of each function called on the `tbl_regression` object  
  > `inputs` Inputs from the `tbl_regression()` function call  
  > `gt_calls` Named list of gt function calls  

`gt_calls` is a named list of saved gt function calls.  The gt calls are run when the object is printed to the console or in an R markdown document. Here's an example of the first few calls saved with `tbl_regression()`:

```{r}
tbl_regression(m1) %>% purrr::pluck("gt_calls") %>% head(n = 3)
```

The gt functions are called in the order they appear, always beginning with the `gt()` function.

If the user does not want a specific gt function to run, any gt call can be omitted in the `as_gt()` function by specifying the `omit` argument.  For example, the `tbl_regression()` call creates many named gt function calls: *`r tbl_regression(m1) %>% purrr::pluck("gt_calls") %>% names() %>% paste(collapse = ", ")`*.  Any one of these can be omitted.  In this example, the default indentation will be omitted from the output.

```{r remedy001}
tbl_regression(m1) %>%
  as_gt(omit = "tab_style_text_indent")
```

## tbl_uvregression

The `tbl_uvregression()` produces a table of univariate regression results.  The function is a wrapper for `tbl_regression()`, and as a result, accepts nearly identical function arguments.  The function's results can be modified in similar ways to `tbl_regression()` and the results reported inline similarly to `tbl_regression()`.

```{r tbl_uvregression}
tbl_uvregression(
 trial,
 method = "glm",
 y = "response",
 method.args = list(family = binomial),
 exponentiate = TRUE,
 pvalue_fun = function(x) style_pvalue(x, digits = 2)
) %>%
  # overrides the default that shows p-values for each level
  add_global() %>%
  # adjusts global p-values for multiple testing (default method: FDR)
  add_q() %>%
  # bold p-values under a given threshold (default 0.05)
  tab_style_bold_p() %>%
  # now bold q-values under the threshold of 0.10
  tab_style_bold_p(t = 0.10, q = TRUE) %>%
  tab_style_bold_labels()
```

