---
title: "Tutorial: tbl_regression"
author: "Daniel D. Sjoberg"
date: "Last Updated: November 29, 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: tbl_regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
library(gtsummary)
library(dplyr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette will walk a reader through the `tbl_regression()` function, and the various functions available to modify and make additions to an existing formatted regression table. `tbl_regression()` uses `broom::tidy()` to perform the initial model formatting, and can accommodate many different model types (e.g. `lm()`, `glm()`, `coxph()`, `survreg()` and more). If your class of model is not supported by `broom::tidy()`, please request support at github.com/ddsjoberg/gtsummary/issues.  In most cases, it is simple to support a new class of model.

To start, a quick note on the `magrittr` package's pipe function, `%>%`.  By default the pipe operator puts whatever is on the left hand side of `%>%` into the first argument of the function on the right hand side.  The pipe function can be used to make the code relating to `tbl_regression()` easier to use, but it is *not* required.  Here are a few examples of how `%>%` translates into typical R notation.

    x %>% f() is equivalent to f(x)
    x %>% f(y) is equivalent to f(x, y)
    y %>% f(x, .) is equivalent to f(x, y)
    z %>% f(x, y, arg = .) is equivalent to f(x, y, arg = z)

## Basic Usage

We will use the `trial` data set throughout this example.  This set contains data from `r nrow(trial)` patients randomized to a new adjuvant treatment or placebo.  The outcome is a binary tumor response.  Each variable in the data frame has been assigned an attribute label (i.e. `attr(trial$trt, "label") == "Treatment Randomization")`

    trt      Treatment Randomization
    age      Age, yrs
    marker   Marker Level, ng/mL
    stage    T Stage
    grade    Grade
    response Tumor Response

The default output from `tbl_regression()` is meant to be publication ready.  Let's start by creating a regression model table from the `trial` data set included in the `gtsummary` package.  We will predict tumor response using age, stage, and grade using a logistic regression model.  

```{r, message=FALSE}
# build logistic regression model
m1 = glm(response ~ age + stage + grade, trial, family = binomial(link = "logit"))

# format results into data frame
tbl_regression(m1, exponentiate = TRUE)
```

Because the variables in the data set were labeled, the labels were carried through into the model regression results table.  (Had the data not been labeled, the default is to display the variable name.)  The model was recognized as logistic regression with coefficients exponentiated, so the header displayed "OR" for odds ratio. The correct reference group has also been added to the table.

Imagine you are not interested in the individual p-values for the stage and grade levels.  You can use the `add_global()` function to add a global p-value for each.  The `add_global()` function is a wrapper for `cars::Anova()`, which calculates LRT and Wald test statistics (a few other methods are also available depending on the type of model). The p-value presented is the default from `cars::Anova()` for your type of model.  You can pass arguments to this function via `add_global()` to change the defaults.

```{r, message=FALSE}
# format results into data frame with global p-values
tbl_regression(m1, exponentiate = TRUE) %>% 
  add_global()
```

## Format Model Results

There are formatting options available, such as bolding and italicizing rows.  In the example below, variable labels are bolded, and the levels of categorical levels are italicized.

```{r, message=FALSE}
# format results into data frame with global p-values
tbl_regression(m1, exponentiate = TRUE) %>% 
  tab_style_bold_labels() %>% 
  tab_style_italicize_levels()
```

The `tbl_regression()` has default methods for rounding p-values and model coefficients.  The default p-value formatting function is `style_pvalue()` and `style_sigfig()` for the coefficients.  In the example below, the large large p-values were rounded to 1 decimal place rather than 2.

```{r, message=FALSE}
# format results into data frame with global p-values
tbl_regression(
  m1, 
  exponentiate = TRUE, 
  pvalue_fun = function(x) style_pvalue(x, digits = 2)
) 
```

##  Report Results Inline (coming soon!)

## Under the Hood

When you print the output from the `tbl_regression()` function into the R console or into an Rmarkdown, there are default printing functions that are called in the background: `print.tbl_regression()` and `knit_print.tbl_regression()`.  The true output from `tbl_regression()` is a named list, but when you print into the R console the interesting portions are displayed from the `.$table_body` data frame.

```{r}
ls(m1 %>% tbl_regression())
```

There is additional information stored in the `tbl_regression()` output list. 

- `table_body` data frame with regression summary statistics
- `model_obj` regression model object
- `n` sample size
- `inputs` Inputs from the function call. Not only is the call stored, but the values of the inputs as well.  

```{r}
print.listof(m1 %>% tbl_regression())
```

## tbl_uvregression

The `tbl_uvregression()` produces a table of univariate regression results.  The function is a wrapper for `tbl_regression()`, and as a result, accepts nearly identical function arguments.  The function's results can be modified in similar ways to `tbl_regression()` as well.

```{r tbl_uvregression}
# rounding pvalues to 2 decimal places, 
# and adding global p-values,
# and bold_labels
tbl_uvregression(
 trial,
 method = "glm",
 y = "response",
 method.args = list(family = binomial),
 exponentiate = TRUE,
 pvalue_fun = function(x) style_pvalue(x, digits = 2)
) %>%
  # overrides the default that shows p-values for each level
  add_global() %>%
  # adjusts global p-values for multiple testing (default method: FDR)
  add_q() %>%
  # bold p-values under a given threshold (default 0.05)
  tab_style_bold_p() %>% 
  # now bold q-values under the threshold of 0.10
  tab_style_bold_p(
    t = 0.10,
    q = TRUE
  ) %>%
  tab_style_bold_labels()
```

