---
title: "Tutorial: tbl_summary"
author: "Daniel D. Sjoberg"
date: "Last Updated: November 29, 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: tbl_summary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(gtsummary)
```

## Introduction

This vignette will walk a reader through the `tbl_summary()` function, and the various functions available to modify and make additions to an existing summary table object.

To start, a quick note on the `magrittr` package's pipe function, `%>%`.  By default the pipe operator puts whatever is on the left hand side of `%>%` into the first argument of the function on the right hand side.  The pipe function can be used to make the code relating to `tbl_summary()` easier to use, but it is *not* required.  Here are a few examples of how `%>%` translates into typical R notation.

    x %>% f() is equivalent to f(x)
    x %>% f(y) is equivalent to f(x, y)
    y %>% f(x, .) is equivalent to f(x, y)
    z %>% f(x, y, arg = .) is equivalent to f(x, y, arg = z)

Here's how this translates into the use of `tbl_summary()`.

    mtcars %>% tbl_summary() is equivalent to tbl_summary(mtcars)
    mtcars %>% tbl_summary(by = "am") is equivalent to tbl_summary(mtcars, by = "am")
    tbl_summary(mtcars, by = "am") %>% add_comparison() is equivalent to
        tbl = tbl_summary(mtcars, by = "am")
        add_comparison(tbl_summary)
    
## Basic Usage

We'll be using the `trial` data set throughout this example.  This set contains data from `r nrow(trial)` patients randomized to a new drug therapy or placebo.  The outcome is a binary tumor response.  Each variable in the data frame has been assigned an attribute label (e.g. `attr(trial$trt, "label") = "Treatment Randomization")`. These labels are displayed in the output table by default.  A data frame without labels will print variable names.

    trt      Treatment Randomization
    age      Age, yrs
    marker   Marker Level, ng/mL
    stage    T Stage
    grade    Grade
    response Tumor Response

```{r, message=FALSE}
library(dplyr)
library(gtsummary)

# printing trial data
head(trial) %>% gt()
```

The default output from `tbl_summary()` is meant to be publication ready.  Let's start by creating a summary statistics table from the `trial` data set built into the gtsummary package.  The `tbl_summary()` can take, minimally, a data set as the only input, and return descriptive statistics for each column in the data frame.

For brevity, keeping a subset of the variables in the trial data set.

```{r, message=FALSE}
trial2 =
  trial %>%
  select(trt, marker, stage)

tbl_summary(trial2)
```

This is a great table, but for trial data the summary statistics should be split by randomization group.  While reporting p-values for baseline characteristics in a randomized trial isn't recommended, we'll do it here as an illustration.  To compare two or more groups, include `add_comparison()` to the function call.

```{r}
tbl_summary(trial2, by = "trt") %>% add_comparison()
```

## Customize Summary Table Output

There are four primary was to customize the output of the summary table.

1. Modify `tbl_summary()` function input arguments
1. Add additional data/information to a summary table with `add_*()` functions
1. Modify summary table appearance with the gtsummary functions
1. Modify table appearance with gt package functions

### Modifying function arguments

The `tbl_summary()` function includes many input options for modifying the appearance.

  > `label` modify the variable labels printed in the table  
  > `statistic` change the summary statistics presented  
  > `digits` number of digits the summary statistics will be rounded to  
  > `missing` whether to display a row with the number of missing observations  

### Functions to add information

The gtsummary package has built-in functions for adding to results from `tbl_summary()`.  The following functions add columns and/or information to the summary table.

  > `add_comparison()` Adds p-values to the output comparing values across groups   
  > `add_overall()` Adds a column with overall summary statistics   
  > `add_n()` Adds a column with N (or N missing) for each variable   
  > `add_stat_label()` Adds a column showing a label for the summary statistics shown in each row   
  > `add_q()` Add a column of q values to control for multiple comparisons   

### gtsummary functions to format table

The gtsummary package comes with functions specifically made to modify and format summary tables.  These functions share the same prefixes as functions from the gt package to ease use.

  > `cols_label_summary()` Relabel columns in summary table using data from the table (i.e. N, percent, etc.)  
  > `tab_style_bold_labels()` Bold variable labels  
  > `tab_style_bold_levels()` Bold variable levels  
  > `tab_style_italicize_labels()` Italicize variable labels  
  > `tab_style_italicize_levels()` Italicize variable levels  
  > `tab_style_bold_p()` Bold significant p-values  

### gt functions to format table

The gt package is packed with many great functions for modifying table output---too many to list here.  Review the package's website for a full listing. https://gt.rstudio.com/index.html

To use the gt package functions with gtsummary tables, the summary table must first be converted into a `gt` object.  To this end, use the `as_gt()` function after modifications have been completed with gtsummary functions.

```
trial %>%
  tbl_summary(by = "trt", missing = "no") %>%
  add_n() %>%
  as_gt() %>%
  <gt functions>
```

### Example

The code below calculates the standard table with summary statistics split by treatment randomization with the following modifications

- Report 'mean (SD)' and 'n / N (\%)'  
- Modify variable labels in the table 
- Use t-test instead of Wilcoxon rank-sum 
- Round large p-values to two decimal place  
- Add column with statistic labels  
- Modify header to include percentages in each group  
- Bold variable labels  
- Italicize variable levels  

```{r}
trial2 %>%
  # build base summary table
  tbl_summary(
    by = "trt",
    # change variable labels
    label = list(
      marker = "Pretreatment Marker Level, ng/mL",
      stage = "Clinical T Stage"
      ),
    # change statistics printed in table
    statistic = list(
      continuous = "{mean} ({sd})",
      categorical = "{n} / {N} ({p}%)"
    )
  ) %>%
  # add p-values to table, perform t-test for the marker,
  # and round large pvalues to two decimal place
  add_comparison(
    test = list(marker = "t.test"),
    pvalue_fun = function(x) style_pvalue(x, digits = 2)
  ) %>%
  # add statistic labels
  add_stat_label() %>%
  # bold variable labels, italicize levels
  tab_style_bold_labels() %>%
  tab_style_italicize_levels() %>%
  # bold p-values under a given threshold (default is 0.05)
  tab_style_bold_p(t = 0.2) %>%
  # include percent in headers
  cols_label_summary(
    stat_by = "**{level}**, N = {n} ({style_percent(p, symbol = TRUE)})"
  )
```

Each of the modification functions have additional options outlined in their respective help files.

##  Report Results Inline (coming soon!)

## Advanced Customization

When you print the output from the `tbl_summary()` function into the R console or into an Rmarkdown, there are default printing functions that are called in the background: `print.tbl_summary()` and `knit_print.tbl_summary()`.  The true output from `tbl_summary()` is a named list, but when you print the object, a formatted version of `.$table_body` is displayed.  All formatting and modifications are made using the `gt` package.

```{r}
tbl_summary(trial2) %>% names()
```

These are the additional data stored in the `tbl_summary()` output list. 

  > `table_body` Data frame with summary statistics  
  > `meta_data` Data frame that is one row per variable with data about each  
  > `by` and `df_by` The by variable name, and a  data frame with information about the by variable  
  > `call_list` Named list of each function called on the `tbl_summary` object  
  > `inputs` Inputs from the `tbl_summary()` function call  
  > `gt_calls` Named list of gt function calls  

`gt_calls` is a named list of saved gt function calls.  The gt calls are run when the object is print to the console or in an R markdown document. Here's an example of the first few calls saved with `tbl_summary()`:

```{r}
tbl_summary(trial2) %>% purrr::pluck("gt_calls") %>% head(n = 3)
```

The gt functions are called in the order they appear, always beginning with the `gt()` function.

If the user does not want a specific gt function to run, any gt call can be omitted in the `as_gt()` function by specifying the `omit` argument.  For example, the `tbl_summary()` call creates many named gt function calls: *`r tbl_summary(trial2) %>% purrr::pluck("gt_calls") %>% names() %>% paste(collapse = ", ")`*.  Any one of these can be omitted.  In this example, the default footnote will be omitted from the output.

```{r remedy001}
tbl_summary(trial2) %>%
  as_gt(omit = "footnote_stat_label")
```
