---
title: "Tutorial: tbl_regression"
author: "Daniel D. Sjoberg"
date: "Last Updated: November 29, 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: tbl_regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette will walk a reader through the `tbl_regression()` function, and the various functions available to modify and make additions to an existing formatted regression table. `tbl_regression()` uses [broom::tidy] to perform the initial model formatting, and can accommodate many different model types (e.g. `lm()`, `glm()`, `survival::coxph()`, `survival::survreg()` and more). If your class of model is not supported , please [request support](https://github.com/ddsjoberg/gtsummary/issues).  In some cases, it is simple to support a new class of model.

To start, a quick note on the magrittr package's pipe function, `%>%`.  By default the pipe operator puts whatever is on the left hand side of `%>%` into the first argument of the function on the right hand side.  The pipe function can be used to make the code relating to `tbl_regression()` easier to use, but it is *not* required.  Here are a few examples of how `%>%` translates into typical R notation.

    x %>% f() is equivalent to f(x)
    x %>% f(y) is equivalent to f(x, y)
    y %>% f(x, .) is equivalent to f(x, y)
    z %>% f(x, y, arg = .) is equivalent to f(x, y, arg = z)

## Setup

Before going through the tutorial, install {gt} and {gtsummary}, and load packages.

```{r, message=FALSE}
if(curl::has_internet()) 
  remotes::install_github("rstudio/gt")
# remotes::install_github("ddsjoberg/gtsummary")

library(gtsummary)
library(dplyr)
```

Examples utilize the {gt} package to generate tables.  The  `knitr::kable()` function will be used to generate tables ff the {gt} package is not available, or if the user requests with `options(gtsummary.print_engine = "kable")`. 

## Basic Usage

We will use the `trial` data set throughout this example.  This set contains data from `r nrow(trial)` patients randomized to a new drug or placebo.  The outcome is a binary tumor response.  Each variable in the data frame has been assigned an attribute label (i.e. `attr(trial$trt, "label") == "Treatment Randomization")`

    trt      Treatment Randomization
    age      Age, yrs
    marker   Marker Level, ng/mL
    stage    T Stage
    grade    Grade
    response Tumor Response

The default output from `tbl_regression()` is meant to be publication ready.  Let's start by creating a regression model table from the `trial` data set included in the {gtsummary} package.  We will predict tumor response using age, stage, and grade using a logistic regression model.  

```{r, message=FALSE}
# build logistic regression model
m1 = glm(response ~ age + stage + grade, trial, family = binomial(link = "logit"))

# format results into data frame
tbl_regression(m1, exponentiate = TRUE)
```

Because the variables in the data set were [labelled](http://larmarange.github.io/labelled/), the labels were carried through into the {gtsummary} output table.  (Had the data not been labelled, the default is to display the variable name.)  The model was recognized as logistic regression with coefficients exponentiated, so the header displayed "OR" for odds ratio. The correct reference group has also been added to the table.

## Customize Output

There are four primary ways to customize the output of the regression model table.

1. Modify `tbl_regression()` function input arguments
1. Add additional data/information to a summary table with `add_*()` functions
1. Modify summary table appearance with the {gtsummary} functions
1. Modify table appearance with {gt} package functions

### Modifying function arguments

The `tbl_regression()` function includes many input options for modifying the appearance.

    label         modify the variable labels printed in the table.  
    exponentiate  exponentiate model coefficients.  
    include       names of variables to include in output. Default is all variables.  
    show_yesno    show both levels of yes/no variables. Default prints on single row.  
    conf.level    confidence level of confidence interval.  
    intercept     logical argument indicates whether to include the intercept in output.  
    estimate_fun  function to round and format coefficient estimates.  
    pvalue_fun    function to round and format p-values.  

### Functions to add information

The {gtsummary} package has built-in functions for adding to results from `tbl_regression()`.  The following functions add columns and/or information to the regression table.

    add_global_p()  adds the global p-value for a categorical variables   
    add_nevent()    adds the number of observed events to the results object    

### {gtsummary} functions to format table

The {gtsummary} package comes with functions specifically made to modify and format summary tables.  These functions share the same prefixes as functions from the {gt} package to ease use.

    bold_labels()       bold variable labels  
    bold_levels()       bold variable levels  
    italicize_labels()  italicize variable labels  
    italicize_levels()  italicize variable levels  
    bold_p()            bold significant p-values  

### {gt} functions to format table

The {gt} package is packed with many great functions for modifying table output---too many to list here.  Review the package's website for a full listing. https://gt.rstudio.com/index.html

To use the {gt} package functions with {gtsummary} tables, the regression table must first be converted into a {gt} object.  To this end, use the `as_gt()` function after modifications have been completed with {gtsummary} functions.

```
m1 %>%
  tbl_regression(exponentiate = TRUE) %>%
  as_gt() %>%
  <gt functions>
```
### Example

There are formatting options available, such as adding bold and italics to text.  In the example below,  
- Variable labels are bold  
- Levels of categorical levels are italicized  
- Global p-values for T Stage and Grade are reported
- P-values less than 0.10 are bold  
- Large p-values are rounded to two decimal places  
- Coefficients are exponentiated to give odds ratios  
- Odds ratios are rounded to 2 or 3 significant figures

```{r, message=FALSE}
# format results into data frame with global p-values
m1 %>%
  tbl_regression(exponentiate = TRUE, 
                 pvalue_fun = function(x) style_pvalue(x, digits = 2),
                 estimate_fun = function(x) style_ratio(x, digits = 3))  %>% 
  add_global_p() %>%
  bold_p(t = 0.10) %>%
  bold_labels() %>% 
  italicize_levels()
```

##  Report Results Inline {#inline_text}

Reproducible reports are an important part of good practices.  We often need to report the results from a table in the text of an R markdown report.  Inline reporting has been made simple with `inline_text()`.

Let's first create a regression model table.

```{r}
tbl_m1 <- tbl_regression(m1, exponentiate = TRUE)
tbl_m1
```

To report the result for `age`, use the following commands inline.

> `` `r
inline_text(tbl_m1, variable = "age")` `` 

Here's how the line will appear in your report.

> `r inline_text(tbl_m1, variable = "age")`

It is reasonable that you'll need to modify the text.  To do this, use the `pattern` argument.  The `pattern` argument syntax follows [glue::glue] format with referenced R objects being inserted between curly brackets.  You have access the to following fields within the `pattern` argument.  The default is `pattern = "{estimate} ({conf.level*100}% CI {conf.low}, {conf.high}; {p.value})"`

    {estimate}   primary estimate (e.g. model coefficient, odds ratio)
    {conf.low}   lower limit of confidence interval
    {conf.high}  upper limit of confidence interval
    {p.value}    p-value
    {conf.level} confidence level of interval
    {N}          number of observations

> Age was not significantly associated with tumor response `` `r
inline_text(tbl_m1, variable = "age", pattern = "(OR {estimate}; 95% CI {conf.low}, {conf.high}; {p.value})")` ``.

> Age was not significantly associated with tumor response `r inline_text(tbl_m1, variable = "age", pattern = "(OR {estimate}; 95% CI {conf.low}, {conf.high}; {p.value})")`.

If you're printing results from a categorical variable, include `level` argument, e.g. `inline_text(tbl_m1, variable = "stage", level = "T3")` resolves to "`r inline_text(tbl_m1, variable = "stage", level = "T3")`".  

The `inline_text` function has arguments for rounding the p-value (`pvalue_fun`) and the coefficients and confidence interval (`estimate_fun`).  These default to the same rounding performed in the table, but can be modified when reporting inline.

## Advanced Customization

When you print the output from the `tbl_regression()` function into the R console or into an R markdown, there are default printing functions that are called in the background: `print.tbl_regression()` and `knit_print.tbl_regression()`.  The true output from `tbl_regression()` is a named list, but when you print the object, a formatted version of `.$table_body` is displayed.  All formatting and modifications are made using the {gt} package.

```{r}
tbl_regression(m1) %>% names()
```

These are the additional data stored in the `tbl_regression()` output list. 

    table_body  data frame with summary statistics  
    n           N included in model  
    model_obj   the model object passed to `tbl_regression`  
    call_list   named list of each function called on the `tbl_regression` object  
    inputs      inputs from the `tbl_regression()` function call  
    gt_calls    named list of {gt} function calls  

`gt_calls` is a named list of saved {gt} function calls.  The {gt} calls are run when the object is printed to the console or in an R markdown document. Here's an example of the first few calls saved with `tbl_regression()`:

```{r}
tbl_regression(m1) %>% purrr::pluck("gt_calls") %>% head(n = 3)
```

The {gt} functions are called in the order they appear, always beginning with the `gt()` function.

If the user does not want a specific {gt} function to run, any {gt} call can be omitted in the `as_gt()` function by specifying the `omit` argument.  For example, the `tbl_regression()` call creates many named {gt} function calls: *`r tbl_regression(m1) %>% purrr::pluck("gt_calls") %>% names() %>% paste(collapse = ", ")`*.  Any one of these can be omitted.  In this example, the default indentation will be omitted from the output.

```{r remedy001}
tbl_regression(m1) %>%
  as_gt(exclude = "tab_style_text_indent")
```

## Univariate Regression {#tbl_uvregression}

The `tbl_uvregression()` produces a table of univariate regression results.  The function is a wrapper for `tbl_regression()`, and as a result, accepts nearly identical function arguments.  The function's results can be modified in similar ways to `tbl_regression()` and the results reported inline similarly to `tbl_regression()`.

```{r tbl_uvregression}
trial %>%
  select(-death, -ttdeath, -stage) %>%
  tbl_uvregression(method = glm,
                   y = response,
                   method.args = list(family = binomial),
                   exponentiate = TRUE,
                   pvalue_fun = function(x) style_pvalue(x, digits = 2)) %>%
  # overrides the default that shows p-values for each level
  add_global_p() %>%
  # adjusts global p-values for multiple testing (default method: FDR)
  add_q() %>%
  # bold p-values under a given threshold (default 0.05)
  bold_p() %>%
  # now bold q-values under the threshold of 0.10
  bold_p(t = 0.10, q = TRUE) %>%
  bold_labels()
```

## Setting Default Options {#options}

The {gtsummary} regression functions and their related functions have sensible defaults for rounding and formatting results.  If you, however, would like to change the defaults there are a few options.  The default options can be changed for in a single script with addition an `options()` command in the script.  The defaults can also be set on the project- or user-level R profile, `.Rprofile`.

```{r, eval=FALSE}
help("Rprofile")

usethis::edit_r_profile()
```

The following parameters are available to be set:

| Description                                         | Example                                                                                   | Functions                                                                        |
|-----------------------------------------------------|-------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------|
| Formatting and rounding p-values                    | `options(gtsummary.pvalue_fun = function(x) style_pvalue(x, digits = 2)`                  | `add_p()`, `tbl_regression()`, `tbl_uvregression()`, `add_global_p()`, `add_q()` |
| Formatting and rounding for regression coefficients | `options(gtsummary.tbl_regression.estimate_fun = function(x) style_sigfig(x, digits = 3)` | `tbl_regression()`, `tbl_uvregression()`                                         |
| Set level for limits                                | `options(gtsummary.conf.level = 0.90)`                                                    | `tbl_regression()`, `tbl_uvregression()`                                         |
| Print tables with `gt` or `kable`                   | `options(gtsummary.print_engine = "kable")`   `options(gtsummary.print_engine = "gt")`    | All `tbl_*()` functions                                                          |

