---
title: "Tutorial: tbl_summary"
author: "Daniel D. Sjoberg"
date: "Last Updated: June 20, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Tutorial: tbl_summary}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette will walk a reader through the `tbl_summary()` function, and the various functions available to modify and make additions to an existing table summary object.

To start, a quick note on the {magrittr} package's pipe function, `%>%`.  By default the pipe operator puts whatever is on the left hand side of `%>%` into the first argument of the function on the right hand side.  The pipe function can be used to make the code relating to `tbl_summary()` easier to use, but it is *not* required.  Here are a few examples of how `%>%` translates into typical R notation.

    x %>% f() is equivalent to f(x)
    x %>% f(y) is equivalent to f(x, y)
    y %>% f(x, .) is equivalent to f(x, y)
    z %>% f(x, y, arg = .) is equivalent to f(x, y, arg = z)

Here's how this translates into the use of `tbl_summary()`.

    mtcars %>% tbl_summary() is equivalent to tbl_summary(mtcars)
    mtcars %>% tbl_summary(by = am) is equivalent to tbl_summary(mtcars, by = am)
    tbl_summary(mtcars, by = am) %>% add_p() is equivalent to
        tbl = tbl_summary(mtcars, by = am)
        add_p(tbl)
    
## Setup

Before going through the tutorial, install {gtsummary} and {gt}.

```{r, include=FALSE}
library(gtsummary)
library(dplyr)

# installing gt 
if (curl::has_internet()) {
  # adding tmpdir to libPath
  temp_path <- file.path(tempdir(), "gt_folder")
  dir.create(temp_path)
  lib_path <-.libPaths()
  .libPaths(c(lib_path, temp_path))
  
  # installing gt
  remotes::install_github("rstudio/gt", lib = temp_path)
}
```

```{r, eval=FALSE}
install.packages("gtsummary")
remotes::install_github("rstudio/gt")

library(gtsummary)
library(dplyr)
```

- If you experience issues installing {gt} on Windows, install [Rtools from CRAN](https://cran.r-project.org/bin/windows/Rtools/), restart R, and attempt installation again.

The vignettes hosted on https://cran.r-project.org may not use the {gt} package to print tables.  For examples with {gt}, browse to the {gtsummary} [website](http://www.danieldsjoberg.com/gtsummary/). The  `knitr::kable()` function is used to generate tables when the {gt} package is not available, or if the user requests with `options(gtsummary.print_engine = "kable")`. 

## Basic Usage

We'll be using the `trial` data set throughout this example.  This set contains data from `r nrow(trial)` patients randomized to a new drug therapy or placebo.  The outcome is a binary tumor response.  Each variable in the data frame has been assigned an attribute label (i.e. `attr(trial$trt, "label") == "Treatment Randomization")` with the [labelled](http://larmarange.github.io/labelled/) package. These labels are displayed in the output table by default.  A data frame without labels will print variable names.

    trt      Treatment Randomization
    age      Age, yrs
    marker   Marker Level, ng/mL
    stage    T Stage
    grade    Grade
    response Tumor Response
    death    Patient Died
    ttdeath  Years from Randomization to Death/Censor

```{r, message=FALSE}
# printing trial data
head(trial) %>% knitr::kable()
```

The default output from `tbl_summary()` is meant to be publication ready.  Let's start by creating a table of summary statistics from the `trial` data set.  The `tbl_summary()` can take, at minimum, a data set as the only input, and returns descriptive statistics for each column in the data frame.

For brevity, keeping a subset of the variables in the trial data set.

```{r, message=FALSE}
trial2 =
  trial %>%
  select(trt, marker, stage)

tbl_summary(trial2)
```

This is a great table, but for trial data the summary statistics should be split by randomization group.  While reporting p-values for baseline characteristics in a randomized trial isn't recommended, we'll do it here as an illustration of the feature.  To compare two or more groups, include `add_p()` with the function call.

```{r}
tbl_summary(trial2, by = trt) %>% add_p()
```

## Customize Output

There are four primary ways to customize the output of the summary table.

1. Modify `tbl_summary()` function input arguments
1. Add additional data/information to a summary table with `add_*()` functions
1. Modify summary table appearance with the {gtsummary} functions
1. Modify table appearance with {gt} package functions

### Modifying function arguments

The `tbl_summary()` function includes many input options for modifying the appearance.

    label       specify the variable labels printed in table  
    type        specify the variable type (e.g. continuous, categorical, etc.)
    statistic   change the summary statistics presented  
    digits      number of digits the summary statistics will be rounded to  
    missing     whether to display a row with the number of missing observations 
    sort        change the sorting of categorical levels by frequency
    percent     print column, row, or cell percentages

### Functions to add information

The {gtsummary} package has built-in functions for adding to results from `tbl_summary()`.  The following functions add columns and/or information to the summary table.

    add_p()           add p-values to the output comparing values across groups   
    add_overall()     add a column with overall summary statistics   
    add_n()           add a column with N (or N missing) for each variable   
    add_stat_label()  add a column showing a label for the summary statistics shown in each row   
    add_q()           add a column of q values to control for multiple comparisons   

### {gtsummary} functions to format table

The {gtsummary} package comes with functions specifically made to modify and format summary tables.  

    modify_header()         relabel columns in summary table  
    bold_labels()           bold variable labels  
    bold_levels()           bold variable levels  
    italicize_labels()      italicize variable labels  
    italicize_levels()      italicize variable levels  
    bold_p()                bold significant p-values  

### {gt} functions to format table

The {gt} package is packed with many great functions for modifying table output---too many to list here.  Review the package's website for a full listing. https://gt.rstudio.com/index.html

To use the {gt} package functions with {gtsummary} tables, the summary table must first be converted into a `gt` object.  To this end, use the `as_gt()` function after modifications have been completed with {gtsummary} functions.

```
trial %>%
  tbl_summary(by = trt, missing = "no") %>%
  add_n() %>%
  as_gt() %>%
  <gt functions>
```

### Example

The code below calculates the standard table with summary statistics split by treatment randomization with the following modifications

- Report 'mean (SD)' and 'n / N (\%)'  
- Round the marker mean and SD to 1 and 2 places, respectively
- Modify variable labels in the table 
- Use t-test instead of Wilcoxon rank-sum 
- Round large p-values to two decimal place  
- Add column with statistic labels  
- Modify header to include percentages in each group  
- Bold variable labels  
- Italicize variable levels  

```{r}
trial2 %>%
  # build base summary table
  tbl_summary(
    by = trt,
    # change variable labels
    label = list(vars(marker) ~ "Marker, ng/mL",
                 vars(stage) ~ "Clinical T Stage"),
    # change statistics printed in table
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} / {N} ({p}%)"),
    digits = list("marker" ~ c(1, 2))
  ) %>%
  # add p-values, report t-test, round large pvalues to two decimal place
  add_p(test = list(vars(marker) ~ "t.test"),
                 pvalue_fun = function(x) style_pvalue(x, digits = 2)) %>%
  # add statistic labels
  add_stat_label() %>%
  # bold variable labels, italicize levels
  bold_labels() %>%
  italicize_levels() %>%
  # bold p-values under a given threshold (default is 0.05)
  bold_p(t = 0.2) %>%
  # include percent in headers
  modify_header(stat_by = "**{level}**, N = {n} ({style_percent(p, symbol = TRUE)})")
```

Each of the modification functions have additional options outlined in their respective help files.

## Select Helpers {#select_helpers}

The {gtsummary} package includes a set of functions meant to help the user specify function arguments for groups of variables.  For example, if all continuous variables will be summarized in `tbl_summary()` as minimum and maximum, the `all_continuous()` function can be used: `statistic = all_continuous() ~ "{min} to {max}"`

The set of select helper function includes the functions in the {tidyselect} package (used throughout the tidyverse), and functions specific to {gtsummary}. There are four types of select helpers.  

1. Functions in the {tidyselect} package used throughout the tidyverse, including `vars()` from the {dplyr} package.  

1. Functions to select groups of variables based on their attributes like class or type.  

1. Functions to select groups of variables based on their display type in `tbl_summary()`  

1. List variables in a vector, e.g. `"age"` or `c("age", "stage")`

| {tidyselect}                                                                                                                 | attribute                                                                                            | summary type                                                 | character vector             |
|------------------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------|--------------------------------------------------------------|------------------------------|
| `starts_with()`, `ends_with()`, `contains()`, `matches()`, `one_of()`, `everything()`, `num_range()`, `last_col()`, `vars()` | `all_numeric()`, `all_integer()`, `all_logical()`, `all_factor()`, `all_character()`, `all_double()` | `all_continuous()`, `all_categorical()`, `all_dichotomous()` | `"age"` or `c("age", "stage")` |

The select helpers are primarily used in `tbl_summary()` and its related functions, e.g. `add_p()`.  We'll review a few examples illustrating their use.

### Examples

In the example below, we report the mean and standard deviation for continuous variables, and percent for all categorical.  We'll report t-tests rather than Wilcoxon rank-sum test for continuous variables, and report Fisher's exact test for response.

Note that dichotomous variables are, by default, included with `all_categorical()`.  Use `all_categorical(dichotomous = FALSE)` to exclude dichotomous variables.

```{r}
trial %>%
  dplyr::select(trt, response, age, stage, marker, grade) %>%
  tbl_summary(
    by = trt,
    type = list(c("response", "grade") ~ "categorical"), # select by variables in a vector
    statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%") # select by summary type/attribute
  ) %>%
  add_p(test = list(contains("response") ~ "fisher.test", # select using functions in tidyselect
                    all_continuous() ~ "t.test"))
```

## Report Results Inline {#inline_text}

Reproducible reports are an import part of good practices.  We often need to report the results from a table in the text of an R markdown report.  Inline reporting has been made simple with `inline_text()`.

First create a basic summary table.

```{r}
tab1 = tbl_summary(trial2, by = trt)
tab1
```

To report the median (IQR) of the marker levels in each group, use the following commands inline.

> The median (IQR) marker level in the drug and placebo groups are `` `r
inline_text(tab1, variable = "marker", column = "Drug")` `` and `` `r
inline_text(tab1, variable = "marker", column = "Placebo")` ``, respectively.

Here's how the line will appear in your report.

> The median (IQR) marker level in the drug and placebo groups are `r inline_text(tab1, variable = "marker", column = "Drug")` and `r inline_text(tab1, variable = "marker", column = "Placebo")`, respectively.

If you display a statistic from a categorical variable, include the `level` argument.  

> `` `r
inline_text(tab1, variable = "stage", level = "T1", column = "Drug")` `` resolves to "`r inline_text(tab1, variable = "stage", level = "T1", column = "Drug")`"

For more detail about inline code, review to the  [RStudio documentation page](https://rmarkdown.rstudio.com/lesson-4.html).

## Advanced Customization {#advanced}

When you print output from the `tbl_summary()` function into the R console or into an R markdown, there are default printing functions that are called in the background: `print.tbl_summary()` and `knit_print.tbl_summary()`.  The true output from `tbl_summary()` is a named list, but when you print the object, a formatted version of `.$table_body` is displayed.  All formatting and modifications are made using the {gt} package.

```{r}
tbl_summary(trial2) %>% names()
```

These are the additional data stored in the `tbl_summary()` output list. 

    table_body   data frame with summary statistics  
    meta_data    data frame that is one row per variable with data about each  
    by, df_by    the by variable name, and a  data frame with information about the by variable  
    call_list    named list of each function called on the `tbl_summary` object  
    inputs       inputs from the `tbl_summary()` function call  
    gt_calls     named list of {gt} function calls  
    kable_calls  named list of function calls to be applied before knitr::kable()  

`gt_calls` is a named list of saved {gt} function calls.  The {gt} calls are run when the object is printed to the console or in an R markdown document. Here's an example of the first few calls saved with `tbl_summary()`:

```{r}
tbl_summary(trial2) %>% purrr::pluck("gt_calls") %>% head(n = 5)
```

The {gt} functions are called in the order they appear, always beginning with the `gt()` function.

If the user does not want a specific {gt} function to run (i.e. would like to change default printing), any {gt} call can be excluded in the `as_gt()` function by specifying the `exclude` argument.  For example, the `tbl_summary()` call creates many named {gt} function calls: *`r tbl_summary(trial2) %>% purrr::pluck("gt_calls") %>% names() %>% paste(collapse = ", ")`*.  Any of these can be excluded.  In the example below, the default footnote will be excluded from the output.

After the `as_gt()` function is run, additional formatting may be added to the table using {gt} formatting functions. In the example below, a spanning header for the `by = ` variable is included with the {gt} function `tab_spanner()`.

```{r as_gt2, eval=FALSE}
tbl_summary(trial2, by = trt) %>%
  as_gt(exclude = "tab_footnote") %>%
  gt::tab_spanner(label = "Randomization Group",
                  columns = gt::starts_with("stat_"))
```

```{r as_gt1, echo=FALSE}
# this code chunk only works if gt is installed
if (requireNamespace("gt", quietly = TRUE)) {
  tbl_summary(trial2, by = trt) %>%
    as_gt(exclude = "tab_footnote") %>%
    gt::tab_spanner(label = "Randomization Group",
                    columns = gt::starts_with("stat_"))
}
```

## Setting Default Options {#options}

The `tbl_summary()` function and it's related functions have sensible defaults for rounding and formatting results.  If you, however, would like to change the defaults there are a few options.  The default options can be changed for a single script with addition an `options()` command in the script.  The defaults can also be set on the project- or user-level R profile, `.Rprofile`.

```{r, eval=FALSE}
help("Rprofile")

usethis::edit_r_profile()
```

The following parameters are available to be set:

```{r, echo=FALSE}
data.frame(
  `Description` = c("Formatting and rounding p-values", 
                    "Formatting and rounding percentages",
                    "Print tables with `gt` or `kable`"),
  `Example` = c("`options(gtsummary.pvalue_fun = function(x) gtsummary::style_pvalue(x, digits = 2))`",
                '`options(gtsummary.tbl_summary.percent_fun = function(x) sprintf("%.2f", 100 * x))`',
                '`options(gtsummary.print_engine = "kable")`   `options(gtsummary.print_engine = "gt")`'),
  `Functions` = c("`add_p()`, `tbl_regression()`, `tbl_uvregression()`",
                  "`tbl_summary()`",
                  "All `tbl_*()` functions")
) %>% 
  knitr::kable()
```

When setting default rounding/formatting functions, set the default to a function object rather than an evaluated function.  For example, if you want to round percentages to 3 significant figures use,

```{r,eval=FALSE}
options(gtsummary.tbl_summary.percent_fun = function(x) sigfig(x, digits = 3))
```

