---
title: "Tutorial: fmt_table1"
author: "Daniel D. Sjoberg"
date: "Last Updated: November 29, 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: fmt_table1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette will walk a reader through the `fmt_table1()` function, and the various functions available to modify and make additions to an existing Table 1.

To start, a quick note on the `magrittr` package's pipe function, `%>%`.  By default the pipe operator puts whatever is on the left hand side of `%>%` into the first argument of the function on the right hand side.  The pipe function can be used to make the code relating to `fmt_table1()` easier to use, but it is *not* required.  Here are a few examples of how `%>%` translates into typical R notation.

    x %>% f() is equivalent to f(x)
    x %>% f(y) is equivalent to f(x, y)
    y %>% f(x, .) is equivalent to f(x, y)
    z %>% f(x, y, arg = .) is equivalent to f(x, y, arg = z)

Here's how this translates into the use of `fmt_table1()`.

    mtcars %>% fmt_table1() is equivalent to fmt_table1(mtcars)
    mtcars %>% fmt_table1(by = "am") is equivalent to fmt_table1(mtcars, by = "am")
    fmt_table1(mtcars, by = "am") %>% add_comparison() is equivalent to
        t = fmt_table1(mtcars, by = "am")
        add_comparison(t)
    
## Basic Usage

We'll be using the `trial` data set throughout this example.  This set contains data from 200 patients randomized to a new adjuvant therapy or placebo.  The outcome is a binary tumor response.  Each variable in the data frame has been assigned an attribute label (i.e. `attr(trial$trt, "label") = "Treatment Randomization")`. These labels are displayed in the output table by default.  A data frame without labels will print variable names.

    trt      Treatment Randomization
    age      Age, yrs
    marker   Marker Level, ng/mL
    stage    T Stage
    grade    Grade
    response Tumor Response

```{r, message=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(gtsummary)

# printing trial data
head(trial) %>% kable()
```

The default output from `fmt_table1()` is meant to be publication ready.  Let's start by creating a descriptive statistics table from the `trial` data set built into the `gtsummary` package.  The `fmt_table1()` can take, minimally, a data set as the only input, and return descriptive statistics for each column in the data frame.

For brevity, keeping a subset of the variables in the trial data set.

```{r, message=FALSE}
trial2 =
  trial %>%
  select(trt, marker, stage)

fmt_table1(trial2)
```

> If your output does not appear in a formatted table, it is likely due to a known issue in the `knitr::kable()` function.  One way around the issue to to add styling from the **kableExtra** package.  
`fmt_table1(trial2) %>% as_tibble() %>% knitr::kable() %>% kableExtra::kable_styling()`

This is a great table, but for trial data the summary statistics should be split by randomization group.  While reporting p-values for a randomized trial isn't recommended, we'll do it here as an illustration.  To compare two or more groups, include `add_comparison()` to the function call.

```{r}
fmt_table1(trial2, by = "trt") %>% add_comparison()
```

## Customize Table 1 Output

It's also possible to add information to `fmt_table1()` output.  The code below calculates the standard table with summary statistics split by treatment randomization with the following modifications

- Report 'mean (SD)' and 'n / N (\%)'  
- Use t-test instead of Wilcoxon rank-sum 
- Do not add row for number of missing observations
- Round large p-values to two decimal place  
- Add column of q-values (p-values adjusted using FDR)
- Add column reporting summary statistics for the cohort overall  
- Add column reporting N not missing for each variable  
- Add column with statistic labels  
- Modify header to include percentages in each group  
- Bold variable labels  
- Italicize variable levels  

```{r}
trial2 %>%
  # build base table 1
  fmt_table1(
    by = "trt",
    # change variable labels
    label = list(
      marker = "Pretreatment Marker Level, ng/mL",
      stage = "Clinical T Stage"
      ),
    # change statistics printed in table
    statistic = list(
      continuous = "{mean} ({sd})",
      categorical = "{n} / {N} ({p}%)"
    ),
    missing = "no"
  ) %>%
  # add p-values to table, perform t-test for the marker,
  # and round large pvalues to two decimal place
  add_comparison(
    test = list(marker = "t.test"),
    pvalue_fun = function(x) fmt_pvalue(x, digits = 2)
  ) %>%
  # add q-values (p-values adjusted for multiple testing)
  add_q(pvalue_fun = function(x) fmt_pvalue(x, digits = 2)) %>%
  # add overall column
  add_overall() %>%
  # add column with N
  add_n() %>%
  # add statistic labels
  add_stat_label() %>%
  # bold variable labels, italicize levels
  bold_labels() %>%
  italicize_levels() %>%
  # bold p-values under a given threshold (default 0.05)
  bold_p(t = 0.2) %>%
  # include percent in headers
  modify_header(
    stat_by = c("{level}", "N = {n} ({p}%)"),
    stat_overall = c("All Patients", "N = {N} (100%)")
  )
```

Each of the modification functions have additional options outlined in their respective help files.

##  Report Results Inline

Having a well formatted and reproducible table is a great!  But we often need to report the results from a table in the text of an Rmarkdown report.  Inline reporting has been made simple with `inline_text()`.

Let's first create a basic Table 1.

```{r}
tab1 = fmt_table1(trial2, by = "trt")
tab1
```

To report the median (IQR) of the marker levels in each group, use the following commands inline.

> The median (IQR) marker level in the drug and placebo groups are `` `r
inline_text(tab1, cell = "marker:Drug")` `` and `` `r
inline_text(tab1, cell = "marker:Placebo")` ``, respectively.

Here's how the line will appear in your report.

> The median (IQR) marker level in the drug and placebo groups are `r inline_text(tab1, cell = "marker:Drug")` and `r inline_text(tab1, cell = "marker:Placebo")`, respectively.

The `cell` argument indicates to `inline_text()` which statistic to display.  Information regarding which statistic to display are separated by `":"`.  The first term indicates the variable name and the last indicates the level of the by variable e.g. `marker:Placebo` would display the summary statistics for the variable `marker` among patients in the `Placebo` group.  If you display a statistic from a categorical variable, include the desired level after the variable name, e.g. `stage:T1:Drug`.  

> `` `r
inline_text(tab1, "stage:T1:Drug")` `` resolves to "`r inline_text(tab1, "stage:T1:Drug")`"


## gtsummary + kableExtra

Need a data frame for any reason (e.g. if you want to get extra fancy with `kableExtra`)? Use generic function `as_tibble` to extract an easy-to-use data frame from any `fmt_table1` object. 

```{r, message=FALSE}
#get data frame from fmt_table1 object
tab1_df <- as_tibble(tab1)
```

If you want to customize anything with `knitr::kable` or [`kableExtra`](https://CRAN.R-project.org/package=kableExtra), you can use the above `as_tibble` along with the function `indent_key` which extracts the row numbers you want indented when knitting your table to HTML. (NOTE: Only load `library(kableExtra)` and use the below if knitting to HTML, this will not work with Word or PDF.)  For more on customizing your tables with `kableExtra` check out the package's [vignette on HTML output](https://haozhu233.github.io/kableExtra/awesome_table_in_html.html).

```{r, message=FALSE}
# knit pretty table
tab1 %>%
  bold_labels() %>% # bold labels in here if you want
  as_tibble() %>%
  kable(
    row.names = FALSE,
    caption = "Table 1: Summary of Patient and Clinical Variables"
  ) %>%
  # Below, using kableExtra functions to do things like change table style, add 
  # grouped column header, footnote, and indent variable categories
  kable_styling(
    bootstrap_options = c("striped", "condensed", "hover"), #popular bootstrap styles
    font_size = 16,
    full_width = FALSE
  ) %>%
  add_header_above(c(" " = 1, "Treatment assignment" = 2)) %>%
  footnote(
    general = "Isn't this footnote so nice?",
    number = c("You can also add numbered or lettered footnotes", "Which is great.")
  ) %>%
  add_indent(indent_key(tab1)) 
```

## Under the Hood

When you print the output from the `fmt_table1()` function into the R console or into an Rmarkdown, there are default printing functions that are called in the background: `print.fmt_table1()` and `knit_print.fmt_table1()`.  The true output from `fmt_table1()` is a named list, but when you print into the R console the interesting portions are displayed from the `.$table1` data frame.

```{r}
t = fmt_table1(trial2, by = "trt") %>% add_comparison()
ls(t)
```

There is additional information stored in the `fmt_table1()` output list. 

- `table1` data frame with summary statistics
- `meta_data` data frame that is one row per variable, and contains information about each variable in the object
- `by` the `by = ` variable name from the function call
- `call` the `fmt_table1` function call
- `call_list` named list of each function called for the `fmt_table1` object. the above example would have two elements in the list: `fmt_table1` and `add_comparison`.  
- `inputs` Inputs from the function call. Not only is the call stored, but the values of the inputs as well.  For example, you can access the data frame passed to `fmt_table1()`.

It is particularly useful to access `.$meta_data` to confirm which statistical tests were used to calculate the p-values in the table.

```{r}
print.listof(t)
```
