---
title: '{gtsummary} benchmark'
author: "Gustavo Zapata Wainberg & Daniel Sjoberg"
output: html_document
---


Updated: `r format(Sys.Date(), "%B %d %Y")`  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(gert)
library(dplyr)
library(stringr)
library(ggplot2)
library(bench) # requires dev version!
library(plotly)
library(purrr)

tryCatch(bench::cb_fetch(remote = "origin"), error = function(e) invisible(NULL))
tryCatch(bench::cb_fetch(remote = "upstream"), error = function(e) invisible(NULL))

# Get the data with previous benchmarks
stored_benchmarks <- bench::cb_read()

# TO REVIEW, THROWS AN ERROR: DS udpated 2021-02-28
# removing duplicate checks within a single HASH (not sure why they are there)
# TODO: Learn why there are duplicates and ensure we're picking the correct one.
# My theory is that there are multiple when I was running the commands
# locally, then pushed to GH where the benchmarks were run again.
stored_benchmarks <-
  stored_benchmarks %>%
  rowwise() %>%
  filter(nrow(benchmarks) > 0) %>% # removing empty results (why are there so many empty results?)
  mutate(
    benchmarks = 
      arrange(benchmarks, file, name, time) %>%
      group_by(file, name) %>%
      slice(n()) %>%
      ungroup() %>%
      list()
  )

# commitInfo <- gert::git_commit_info() # Current commit (?).
# commitInfo$id # ID of current commit for current brach

# This retrieves hashes for all the commits which have a tag (just releases (?))
# taglists <- gert::git_tag_list()

# How to get a tash for a specific version
# hash136 <- taglists[str_detect(taglists$name, "1.3.6"), "commit"]

# Keep a list with just the benchmark outputs
benchmarks_outputs <- stored_benchmarks[["benchmarks"]]

# In order to bind and identify all dataframes within benchmark_outputs
# the list needs to be named, so I'm naming it with the hashs
# Then this argument will be used bind rows
commit_hash <- stored_benchmarks[["commit_hash"]]

# Name list elements
names(benchmarks_outputs) <- commit_hash

# bind rows and add .id
benchmarks_df <- dplyr::bind_rows(benchmarks_outputs, .id = "hash")
```

<!-- This css stuff is just to enable rmarkdown to use a wider area of screen -->

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

## Execution times



```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplotly(
  benchmarks_df  %>% 
           ggplot(aes(x = time, y = mean))+
           geom_point(aes(color = hash))+
           facet_wrap(vars(name), 
                      nrow = length(unique(benchmarks_df$name)), 
                      scales = "free_y", 
                      strip.position = "right")+ # Plotly doesn't understand this
           theme_minimal()+
           theme(legend.position='none')+
           labs(title = "Mean execution time for each function (time per iteration in seconds)",
                subtitle = "seconds", # Plotly doesn't understand subtitles
                y = "",
                x = ""),
         width = 900
)

ggplotly(
  benchmarks_df  %>% 
    mutate(hash_abr = stringr::str_sub(string = hash, start = 1, end = 8)) %>% 
           ggplot(aes(x = hash_abr, y = mean))+
           geom_point(aes(color = time))+
           facet_wrap(vars(name), 
                      nrow = length(unique(benchmarks_df$name)), 
                      scales = "free_y", 
                      strip.position = "right")+ # Plotly doesn't understand this
           theme_minimal()+
           theme(legend.position='none',
                 axis.text.x = element_text(angle = 45, vjust = 1, hjust=0))+
           labs(title = "Mean execution time for each function (time per iteration in seconds)",
                subtitle = "seconds", # Plotly doesn't understand subtitles
                y = "",
                x = ""),
         width = 900
)

```

  
The functions used for the benchmark are:

  `simple = tbl_summary(trial)`  
  
  `complex = tbl_summary(trial, by = trt) %>% add_overall() %>% add_p() %>% add_q(quiet = TRUE) %>% add_n()`

  `big_data = big_trial %>% select(age, grade, trt) %>% tbl_summary(by = trt, missing = "no") %>% add_p()` 
  (NOTE: `big_trial` is a 5000-fold copy of `gtsummary::trial`)
