---
title: "gtsummary + Shiny"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Any gtsummary table can be placed into a [Shiny application](https://shiny.rstudio.com/).
The trick is that the gtsummary must first be converted into a gt table using the `gtsummary::as_gt()` function.
Once the table is a proper gt table, we can use the methods from the gt package to place the table in the Shiny app.
Read more about creating Shiny applications with gt tables here: [https://gt.rstudio.com/reference/gt_output.html](https://gt.rstudio.com/reference/gt_output.html).

# Shiny Example

Below is an example of a Shiny app using the gtsummary package to provide dynamic patient characteristic summaries.

<!-- The email address associated with this shiny app is 'danield.sjoberg+gtsummary@gmail.com' -->
```{=html}
<iframe src="https://gtsummary.shinyapps.io/gtsummary-shiny-example/" width="100%" height="400" frameborder="no"></iframe>
```

If the Shiny app has run out of its free server usage, see below for instructions on running the app locally on your machine.

# Shiny Code

```{r, eval=FALSE, file='https://raw.githubusercontent.com/ddsjoberg/gtsummary-shiny-example/main/app.R'}
```

To run this Shiny app locally on your machine, save the script above as `app.R` in a R Project, open in RStudio, and run the application.
